---
title: "CAVESIM Summary and Status"
author: "Paul Leiby"
date: "August 23, 2018"
output:
  word_document:
    df_print: kable
    fig_caption: yes
    fig_height: 5
    fig_width: 7
    reference_docx: mystyles.docx
    toc: yes
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
  html:
    number_sections: yes
    toc: yes
  pdf_document:
    number_sections: yes
    toc: yes
  powerpoint_presentation:
    reference_doc: ORNL-template.pptx
---

```{r setup, include=FALSE}
# output:
#   beamer_presentation:
#     theme: "AnnArbor"
#     colortheme: "dolphin"
#     fonttheme: "structurebold"
knitr::opts_chunk$set(echo = FALSE)
```

## CAVESIM Status by End FY18

- Q3: Preliminary design of HDV & shared mobility model components (ORNL)
- Q4: Model implementation and runs including multiyear dynamics and new components for LDVs (ORNL)
	- link the model from a static annual to simple dynamic
		- Seek penetration scenario from MA3T or ADOPT/Workflow Taskforce
	- include available cost information for alternative technologies
	- include TNC (ride sharing) to distinguish PKT from VKT
	- (phase 1) benchmark to inputs and results from larger simulations
		- Following the "Workflow" development to understand and ID major features 
	- FY18 Product will be a paper (draft)
    - Feeds Quad Chart item "Annual: Reports on national aggregate energy and travel impacts depending on incentives and economic drivers""

## Base inputs from "Scenario Generation"    -    Technology Info
- Technology set (vehicle types) and mix
- Automation Tech Cost information
- Impacts on energy use
- Impacts on speed and VKT
- CACC and platooning tech and impact assumptions
	- (from POLARIS study initial results)
	- Impacts on energy use
	- Impacts on congestion (ave speed in congested areas)
- TNC tech and impact assumptions
	- (from BEAM-centered work?)
	- Impacts on trip numbers, and length
		- deadheading/repositioning inflation factor on VKT
		- ride pooling reduction factor on VKT per PKT
		- ride-pooling inflation factor on PKT (actual including detours vs. desired trip duration)
			
## Base inputs from "Scenario Generation" -    Demand Response Info
- From WholeTraveler
	- VoTT distributions
		- By travel type: region, roadtype, conditions (congested)
	- Other expectations and behavior, e.g. re TNC 
	- Travel utility functional form
- From Land-Use model
	- Land-use-based rebound effects on PKT demand


## Planned Next Steps for CAVESIM (FY19)

- Revise/update all key parameters from draft values to values consistent with SMART Workflow data and scenarios
- Update additional cost and economic conditions (response elasticities) based on Lit review (share insights and conclusions re parameters with others)
- Bullet 3



## CAVESIM Table Main Inputs (1)
CAVESIM Main Inputs	|Units
-------------------------------------------|---------------------
Reference VKT by year(1), sector(2), Region(6) |	veh-km/yr
Reference PKT by year	                       | pass-km/yr
Energy consumption by fuel type(3) by year, sector(2)	|quad or EJ/yr
Vehicle stock by year, by powertrain(4), control-type(5) | share
Efficiency gains for key technology/mechanisms(7), by powertrain | % reduction in energy intensity, mJ/v-km per % adoption (a constant, or a function)

## CAVESIM Table Main Inputs (2)
CAVESIM Main Inputs	|Units
-------------------------------------------|---------------------
Effect of (regional) congestion on ave speed	| % reduction in speed ave speed, (km/hr) as fn of VKT
Effect of (regional) congestion on energy use	| % increase in energy intensity, (mJ/v-km) as fn of VKT
VoTT by sector, Control-type, and trip purpose	| $/pass-hr
Incentives & Costs: vehicle						| $/veh or $/veh-km
Incentives & Costs: fuel						| $/mJ or $/gge
Incentives & Costs: road use (by region)		| $/veh-km
Incentives & Costs: ride pooling				| $/pass-km

## CAVESIM Table Main Inputs (3)
CAVESIM Main Inputs	|Units
-------------------------------------------|---------------------
Costs of TNC operation by control-type			| $/veh-km
User-cost (disutility) of ride-pooling			| $/pass-km pooled with stranger
Ref Extent of ride-pooling by yr, region	| % of PKT
Deadhead km from car hailing			| ave % incr. VKT/trip
Incremental veh km (reduction) from ride pooling	| ave % decr. VKT/trip
Incremental passenger km travel from ride pooling	| ave % incr. PKTtrip
Cost responses:                         |  
Elasticities of travel w.r.t. full cost	| unitless
Elasticities of veh fuel economy to fuel cost	| unitless
Traveler Utility functions (trading off trip attributes, and travel vs other activities)	| [functional form, or marginal utilities for various attributes]

## CAVESIM Table Main Inputs (Notes)

*NOTES*:
 (1) Outputs at 5 year increments, but can be annual	
 (2) Sector: Passenger road travel; Road freight	
 (3) Fueltype: petroleum, electricity	
 (4) Powertrain: ICE, [HEV?], BEV	
 (5) Control: Manual, AV (for case-dependent AV level)	
 (6) Region: Urban, Highway	
 (7) Mechanisms (currently include): Platooning; Eco-Driving; Traffic-flow Control; Higher-highway Speeds; De-emphasized performance; Accident reduction	

## CAVESIM Table Main Outputs

CAVESIM Main Outputs	| Units
-------------------------------------------|----------------
Travel demand response: VKT by year(1), sector(2)	| veh-km/yr
Travel demand response: PKT by year, passenger road travel	| pass-km/yr
Energy consumption by fuel type(3), year, sector(2)	| quad or EJ/yr
Net average energy intensity for road travel by sector(2), region (6)	| mJ/v-km
Average travel speed by region(6)	| km/hr



## Issues/Questions
- Q: does abstracting from regional/temporal detail make sense?
	- What questions are we trying to answer?
	- How accurate a representation is "good enough to be useful"?
	- What can be learned with such a compact/aggregate system that is hard to learn with the full representation?
		- What about the NREL "Scenario Generation" model (YASDM)
- Q: How t=do we best support/contribute to the MEP measure?


## Static Estimates of Energy Intensity Effects Updated and Made Dynamic

```{r cars, echo = F}
library(knitr)
load(file="CAVdecom.sav")
kable(x = IM, 
      caption= "Intensity Multipliers by Vehicle-type, Year, and Case")
```


```{r pressure, include=F}

## Slide with Plot
plot(pressure)
```