---
title: "CAVESIM Technical Documentation"
author:
  affiliation: Oak Ridge National Laboratory
  email: leibypn@ornl.gov
  name: Paul N. Leiby
date: "Rev. 2019-09-14, Printed `r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Purpose:
------------------

This file describes some of the data structures and functions in CAVESIM (CAVSimV10.Rmd), for internal purposes.

Note: It should be executed after `CAVdecom.Rmd`, so that the data structures are defined and can be summarized.

Data Structures
------------------

### Listing of Case-related Input and Output Dataframes

```{r def_varNameAndIndices_fn, include=FALSE}
varNameAndIndices <- function(x = FALSE) {
  # display the variable (dataframe) name and its indices
  #   See https://stackoverflow.com/questions/10520772/in-r-how-to-get-an-objects-name-after-it-is-sent-to-a-function
  #   Default param value FALSE just assures return for empty call w/o x.
  paste0(deparse(substitute(x)), "[", paste(names(x), collapse = ", "),"]")
}

```

#### 3. Estimating Energy Impacts

These are the primary tables for this section:
```{r, eval=FALSE}
cat("-", varNameAndIndices(IM), "\n")
cat("-", varNameAndIndices(msa), "\n")
cat("-", varNameAndIndices(scenAssumps), "\n")
cat("-", varNameAndIndices(EnergyIntensityChanges), "\n")
cat("-", varNameAndIndices(NetEnergyIntensityChangeTable), "\n")

```

- IM[VC, Year, Mech, Opt, Mid, Pess, Zero] 
    - Intensity Multipliers $mu_{vtks}$ by Vehicle-type, Year, Mechanism and EffectCase or Sensitivity Case

- msa[TechScen, VC, Year, Mech, EffectCase, IM]
    - Intensity Multipliers for mechanism, based on EffectCase associated with each TechScen
    - EffectCase in {Zero, Pess, Mid, Opt}

- scenAssumps[VC, Mech, Year, 1, 2, 3, 4, 5, 6, 7] 
    - Effect Sensitivity Cases by Vehicle Class, Year, AV Mechanism and Tech Scenario
    - This table governs the Tech-scenario-based selection of each mechanism effects, by specifing the EffectCase for each mechanism from the set of potential mechanism effects.

- EnergyIntensityChanges[Year, VC, TechScen, NIE]
    - NIE change $\Delta I_{jtv}$ is product of IM (Intensity Multiplier) over Mechanisms (for TechScen applicable EffectCase)

- NetEnergyIntensityChangeTable[TechScen, HDV_2035, HDV_2050, LDV_2035, LDV_2050] 
    - TechScen summary table for display, bar plot, and comparison with prior test values only.

#### 4. Demand Response to CAVs

These are the primary tables for this section:
```{r, eval=FALSE}
cat("-", varNameAndIndices(DemRespParamNotes), "\n")
cat("-", varNameAndIndices(DemRespParams), "\n")
cat("-", varNameAndIndices(DemRespParamsm), "\n")
cat("-", varNameAndIndices(TravTimeCostBasePars), "\n")
cat("-", varNameAndIndices(VTCostBase), "\n")
cat("-", varNameAndIndices(VTCShrBase), "\n")
cat("-", varNameAndIndices(VTCShrAlt), "\n")
cat("-", varNameAndIndices(RelTotCost), "\n")
cat("-", varNameAndIndices(VMTIncrease), "\n")

cat("-", varNameAndIndices(VMTIncreaseCurrCase), "\n")
cat("-", varNameAndIndices(), "\n")
```

- ElasVKT - input parameter, elasticity of travel demand w.r.t. full (generalized) cost of travel
- ElasPKT - input parameter, elasticity of Passenger travel demand w.r.t. full (generalized) cost of travel
    - extract the applicable ElasPKT from ElasVKT for (non-pooling) demand scenarios

- DemRespParamNotes[DemParam, DemParamNote]
    - descriptive notes on demand response parameters
- DemRespParams[VC, Parameter, Zero, Low, Med, High]
    - Key Parameters for LDV Demand Response (for all Demand Scenarios)
- DemRespParamsm[Parameter, VC, EffectCase, value]
    - Key Parameters for LDV Demand Response (for all Demand Scenarios), by EffectCase, long form.
    - **Note: Energy Intensity change is dependent on the Technology Scenario, and year of interest.**
    - EffectCase in {Zero, Low, Med, High}
    - Parameter in

      ElasVKT          : Elas of travel w.r.t generalized cost,
      ExclVehCapCost   : Exclude Veh Capital Cost (wear & ownership costs) (Choose 0 or 1),
      C_deltaMaint     : Fractional Incr cost from automation (Maintenance, cost/mi),
      C_deltaInsur     : Fractional Incr cost from automation (Insurance, cost/mi),
      C_deltaCapCost   : Fractional Incr cost from automation (vehicle Capital Cost, wear and ownership cost/mi),
      C_deltaTolls     : Fractional Incr cost from automation (Tolls, cost/mi),
      C_deltaPrkng     : Fractional Incr cost from automation (Parking, cost/mi),
      C_deltaRegis     : Fractional Incr cost from automation (Registration, cost/mi),
      C_deltaVoTT      : Fractional Incr cost from automation (VoTT cost/hr),
      ShrECostInt      : Share of energy effic cost gains internalized (visible) to driver (fraction, 0 to 1),
      I_deltaCAV       : Fuel energy cost reduction (from effic, fraction, computed from intensity mechanisms)

#### 4.2 Establish Base Travel Costs for Each Cost Component

- Notes on scenario calculation:
    - Energy Intensity change is dependent on the Technology Scenario, and year of interest. 
    - Demand response is a function of full (generalized) travel cost, and energy intensity affects the energy component of travel cost.5 The full scenario requires updating Demand Response parameters with energy intensity reduction (by vehicle class and year) for this Technology Scenario and year, and updating other costs (travel time cost, accident/insurance costs, vehicle capital and maintenance costs) based on the Demand Scenario assumptions.
    - Unit travel time cost per mile $\underline c_{T,v}$ is differentiated by vehicle class, and depends on average speed and hourly time cost.

- TravTimeCostBasePars[var, V_Class, local, intercity, average, Units, Source_Notes]  
    - "Base Travel Time Cost Parameters"
- VTCostBase[DemScen, VehType, CostCat, VTCost_cpm]
    - "Base Vehicle Travel Cost: Components Share by Vehicle Type, Year=2018" 
        - (VehType in {LDVAvgSedan, LDVAvgLtTruck, HDVClass8})
    - Note: these values are the average of local and intercity.
    - Note: In these original (base) vehicle cost component data, VehType differentiates between 2 classes of LDV (car & light truck)
- VTCShrBase[VehType, CostCat, VTCost_shr] 
    - "Base Vehicle Travel Cost: Components Share by VehType, Year=2018", $\sigma_{vit}^B$ for each cost component *i*

#### 4.4 Calculate Adjusted Travel Cost Components for Demand Scenario Conditions

- VTCShrAlt[DemScen, VehType, VC, CostCat, VTCost_shr]
    - "Table: Alt Travel Cost Components Relative to Base, by Demand Scenario & Vehicle Type"
    - relative cost "shares", $\sigma_{vitdj}$, or actually $\sigma_{dv,vc,i,(tj)}$ because *t* and *j* are implicit, 
    - reflect relative cost adjustments in DemRespParams, need not sum to 1.0
- RelTotCost[DemScen, VehType, VC, Case, Total, ElasVKT, fracVMTIncr]
    - cost components aggregated to total (relative to Base MV), and fractVMTIncr base on ElasVKT w.r.t. RelTotCost
    - results by VehType

#### 4.5 Fractional VMT Changes in CAV Demand Scenario (Single Tech Case)

- VMTIncrease[VC, DemScen, ElasCase, fracVMTIncr, CombScen, USOption, RidePoolAdj, USDmult]
    - fracVMTIncr here is the cost-elastic value from RelTotCost, adjusted by exogenous Underserved Demand or Pooling effects
    - Fractional VMT increase for LDV vehicles is currently taken as an unweighted mean for Sedans and LtTrucks.
- VMTIncreaseCurrCase[VC, DemScen, ElasCase, fracVMTIncr, CombScen, USOption, RidePoolAdj, USDmult]
    - this is the specific set of Combined scenarios wanted with the chosen ElasCase, for comparison with test values

Relative total cost $c_{rel,vtdj}$
The fractional increase in VKT $m_{vtdj}(c_{rel,vtdj})$

#### 5. Policy and Combined Scenario Calculations

These are the primary tables for this section:
```{r, eval=FALSE}
cat("-", varNameAndIndices(CaseEffectsEI), "\n")
cat("-", varNameAndIndices(CaseEffects), "\n")
cat("-", varNameAndIndices(CombScen_Combinations), "\n")
cat("-", varNameAndIndices(CaseAssumps),"\n")
cat("-", varNameAndIndices(CaseWantedEffects), "\n")
cat("-", varNameAndIndices(CaseWantedAssumps), "\n")

```

- CaseEffectsEI[TechScen, Year, VC, NIE]
    - all (exog) vehicle energy intensity changes for all values of TechScen, Year, and Vehicle Class
- CaseEffects[CombScen, DemScen, VC, ElasCase, fracVMTIncr, CombScen_Num, DemScen_Num, TechScen, Year, NIE, EnergyUse]
    - basic results from running many combinations of TechScen, DemSend
    - Binding of results across cases
    - Includes [7 DemScen] x [7 TechScen] x [2 ElasCase] x [2 VC] x [1 Year]
    - Primary results are NIE, fracVMTIncr, EnergyUse
    - EnergyUse = (1+NIE)*(1+fracVMTIncr) - 1.0

Count of cases in CaseEffects for each Demand Scenario and TechScenario:
```{r, eval=FALSE}
table(CaseEffects$DemScen, CaseEffects$TechScen)

```

        1 2 3 4 5 6 7
    DS1 4 4 4 4 4 4 4
    DS2 4 4 4 4 4 4 4
    DS3 4 4 4 4 4 4 4
    DS4 4 4 4 4 4 4 4
    DS5 4 4 4 4 4 4 4
    DS6 4 4 4 4 4 4 4
    DS7 4 4 4 4 4 4 4

```{r, eval=FALSE}
table(CaseEffects$VC, CaseEffects$ElasCase)
```

        High Low
    HDV   49  49
    LDV   49  49
  
- CaseAssumps[CombScen, DemScen, ElasCase, TechScen, VC, Year, fracVMTIncr, NIE, EnergyUse, C_Fuel, C_Time, C_VNonF, C_Total, ElasVKT, USOption, RidePoolAdj, USDmult, Congestion_mitigation, De_emphasized_performance, Eco_driving, Higher_highway_speeds, Improved_crash_avoidance, Increased_feature_load, Platooning, Right_sizing]
  - Same cases and result variables as CaseEffects, with 15 more variables documenting case assumptions
  - Note: sum(CaseAssumps[names(CaseEffects)]!=CaseEffects) ==0

- CombScen_Combinations[CombScen, Year, ElasCase, VC, TechScen, DemScen_Num, Notes] 
    - specifies, for each CombScen and VC: combined cases wanted (TechScen, DemScen, ElasCase, Year)

- CaseWantedEffects[CombScen, DemScen, ElasCase, TechScen, VC, Year, fracVMTIncr, NIE, EnergyUse] 
    - selects the combined scenarios in CaseEffects that match the ones desribed in - CombScen_Combinations by ["Year", "ElasCase", "VC", "TechScen", "DemScen_Num"]
    - this is 12 of 198 combined cases

- CaseWantedAssumps[CombScen, DemScen, ElasCase, TechScen, VC, Year, fracVMTIncr, NIE, EnergyUse, C_Fuel, C_Time, C_VNonF, C_Total, ElasVKT, USOption, RidePoolAdj, USDmult, Congestion_mitigation, De_emphasized_performance, Eco_driving, Higher_highway_speeds, Improved_crash_avoidance, Increased_feature_load, Platooning, Right_sizing] 
    - Same thing as CaseWantedEffects, but including all relevant assumptions

Functions
---------------------
- plotStackedBarShares <- function(df, ds, ts = "TS4") 
    - function to take dataframe of shares by demand scenario and VehType, and
        - for specified demand scenario ds,
        - plot stacked bar graph with one column for each VehType.
- plotStackedBarCostCompons <- function(df, ds, ts = "TS4")
    - function to take dataframe of shares by demand scenario and VehType, and
        - for specified demand scenario ds,
        - plot stacked bar graph with one column for each VehType.

- CombScenResultsForOneTechScen <- function(CurrTS, CurrYr)
  - For a single TechScen, CurrTS, and year CurrYr,
    compute combined scenario outcomes for combinations with all 
      - demand scenarios DemScen
      - elasticity cases ElasCase
      - vehicle clases VC
  - combines TechScen and other inputs needed to compute primary scenario results
      - NIE, 
      - fracVMTIncr (based on elastic response to cost)
      - EnergyUse

- GatherScenInputsForOneTechScen <- function(CurrTS, CurrYr) {

    # For a single TechScen CurrTS, and year CurrYr,
    # gather combined scenario input params for combinations with all
    #   - demand scenarios DemScen
    #   - elasticity cases ElasCase
    #   - vehicle clases VC
    # Select energy intensity changes for the Current Technology Scenario and Year, for both vehicle classes

- percentMPGdecreaseTHWH2014 <- function(S, refMPG = 25.0)
    - MPG vs. Speed in mph - Thomas et al. Approach
    
    
#### 8. Examination of VMT and PMT Given Various Levels of Pooling

These are the primary tables for this section:

```{r, eval=FALSE}
cat("-", varNameAndIndices(scenData), "\n")
cat("-", varNameAndIndices(scenCostData), "\n")
cat("-", varNameAndIndices(occupancyEffects), "\n")
cat("-", varNameAndIndices(occupancyTravelEffects),"\n")
cat("-", varNameAndIndices(AltVMT), "\n")
cat("-", varNameAndIndices(), "\n")
```

- scenData[DemScen, VehType, AccAndIns, C_F, Maintenance, Parking, Registration, C_T, TollsFees, C_Total, C_K, C_H, C_V] 
- scenCostData[DemScen, C_F, C_H, C_T, C_V] 
- occupancyEffects[DemScen, C_F, C_H, C_T, C_V, o, C_Vpmt, rho, C_Tpmt, Pi, C_Tdph, C_Mpmt, PMT, VMT, EnergyUse, VMT_direct, TourContraction, RideDetourExpansion] 
- occupancyTravelEffects[DemScen, o, travelEffect, value] 
- AltVMT[DemScen, o, VMT, PMT, EnergyUse, ElasTourLengthWithPooling] 
    - "Energy Use Relative to Base (DS3), by Occupancy and Route Length Elasticity"
    

-----------------------------

- Next steps for CombScenResultsForOneTechScen() and GatherScenInputsForOneTechScen()
    - use _local_ vars to make iteratively make calculations of many dataframes
        - nie
        - DemRespParams (need DemRespParamsBase)
        - VTCShrAlt (needs VTCShrBase)
        - RelTotCost (needs VTCShrAlt)
        - VMTIncrease (needs) RelTotCost)
        - VMTIncrease
    - collapse the two long, largely duplicative, functions into 1, or have the second call the first


---------------------------------------

## Disaggregating Energy Intensity and Demand responses by subcategory

### Vehicle/Travel Categories

We have the following subcategories of passenger road vehicle travel

- r: Region or Roadtype, $r \in \{ Arterial, Highway \}$
- c: Class of vehicle, $c in \{ Car, LtTruck, HDV \}$
- A: Automation=, $A \in \{ None, Partial, Full \}$
- F: Fuel, $F \in \{ Gas, Elec \}$
- U: Use (Ownership), $U \in \{ Private, Shared \}$

Thus vehicle type, vehicle energy intensity *I*, travel demand (miles) *D* and energy use *E* are all indexed, or subscripted by $r,c,a,f,u$ as well as year $y$

#### Energy as Product of Vehicle Energy Intensity and VMT

Demand scenarios are largely specified by the fraction of vehicle stock, or fraction of trips satisfied, by vehicles of eack type $K_{y,r,c,a,f,u}$.

The energy use by each vehicle type is determined by its intensity and travel demand (VMT or $D$):

$$E_{y,r,c,a,f,u} = I_{y,r,c,a,f,u} \cdot D_{y,r,c,a,f,u}$$

#### Vehicle Energy Intensity 

For the full CAV penetration scenarios (`DemScen == Test`), without electrification, we computed fuel intensity relative to the base vehicle travel (average Region/Road)
$$\rho^I_{y,r,c,a,f,u} \equiv I_{y,r,c,a,f,u} /I_{c,base}$$
The `Test` scenario results are based on year and vehicle Class, and apply to a fully automated vehicle (), gasoline drivetrain ($F = \text{gas}$), in average private (non-shared) use ($U = \text{shared}$):

$$I_{y,r=ave,c,a=F,f=G,u=P}  = I_{y,c,base} \cdot (1+\Delta I_{y,c})$$

$$\rho^I_{y,c} \equiv I_{y,r=ave,c,a=F,f=G,u=P} /I_{y,c,base} = 1+\Delta I_{y,c}$$

Where 
$$I_{c,base} = I_{y, r=ave, c, a=None, f=Gas, u=Private} $$

In the current approach, road travel activity (travel demand) is disaggregated by the vehicle type (Class, Fuel, Automation) and demand segment (Roadtype and Use case, private or shared). Each vehicle's energy intensity is based on those categories,  which alter the energy intensity of the base vehicle type (conventional gasoline, no automation). That is

$$I_{y,r,c,a,f,u}  = I_{y,base} \cdot (1+\Delta I_{r})(1+\Delta I_{c})(1+\Delta I_{a})(1+\Delta I_{f,a})(1+\Delta I_{u})$$
$$I_{y,r,c,a,f,u}  = I_{y,base} \cdot (m^I_{r})(m^I_{c})(m^I_{a})(m^I_{fa})(m^I_{u})$$
The intensity adjustment factors $m^I_{.}$ also depend to some degree on the vehicle/driving categories.

Some of these intensity effect modifiers apply to the incremental effect of automation on vehicle energy intensity. Others, like `fuel_type`, apply also to the base energy intensity of the vehicle, that is, reflecting its electrification.

#### Similarly, for Travel:

Travel demand is similarly disaggregated


fracVMTIncr = (1 + fracVMTIncr) * USDmult * (1 + RidePoolAdjFrac) - 1

$$D_{y,r,c,a,f,u} = D_{y,r,c,a,f,u}$$

#### With combined effect on Energy use
$$E_{y,r,c,a,f,u} = I_{y,r,c,a,f,u} \cdot D_{y,r,c,a,f,u}$$

$$\rho^E_{y,r,c,a,f,u} \equiv \frac {E_{y,r,c,a,f,u}} {E_{y,r,c,a,f,u}}$$

$$\rho^E_{y,r,c,a,f,u} = \frac {I_{y,r,c,a,f,u} \cdot D_{y,r,c,a,f,u}} {I_{y,r,c,a,f,u} \cdot D_{y,r,c,a,f,u}}$$

Note: some important parameters are correctly indexed by these vehicle/travel types:

- VoTT depend on *r* region, *a, u* (not *f*)
- Vehicle non-energy operating costs depends on Vehicle Class *c, a, f*
- Vehicle energy intensity depends on *c, r, a, f,* (but not *u*)



