---
title: "Elas VKT and Elas PKT"
author: "Paul N. Leiby"
date: "3/27/2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=F, message=F, warning=F}
rm(list=ls())
library(tidyverse)
library(knitr)
echoWorking = F
```

```{r basicUSTripStats, echo=echoWorking}
# U.S. DOT-NHTSA 2018. Summary of Travel Trends 2017 National Household Travel Survey - 2017_nhts_summary_travel_trends. https://nhts.ornl.gov/assets/2017_nhts_summary_travel_trends.pdf
# U.S. DOT-BTS 2019. Tranportation Statistics Annual Report - Final - TSAR-Full-2018-Web-Final, Feb. 2019 https://www.bts.gov/sites/bts.dot.gov/files/docs/browse-statistical-products-and-data/transportation-statistics-annual-reports/TSAR-Full-2018-Web-Final.pdf
numHHvehicles_US2017 = 223E6  # US DOT-NHTSA 2018 p. 68
numAveTripsPerPersonDaily_US2017 = 3.37  # US DOT-NHTSA 2018 p. 40
numAveTripsPerPersonAnnual_US2017 = 1231  # US DOT-NHTSA 2018 p. 32
numAvePersonTripsAllModesAnnual_US2017 = (64582+6048+142754+40303+102327+15139)*1E6  # US DOT-NHTSA 2018 p. 29
numAvePersonTripsPrivateVehicleAnnual_US2017 = 0.826 * numAvePersonTripsAllModesAnnual_US2017  # US DOT-NHTSA 2018 p. 30
totalRegisteredVehicles_US2016 = 268799083 # US DOT-BTS 2018 p. 1-4 (HH and non-HH?)
numAveVehiclesPerHH_US2017 = 1.88  # US DOT-NHTSA 2018 p. 59
numHHUS_2017 = numHHvehicles_US2017/numAveVehiclesPerHH_US2017  # lower estimate than FRED Total Household data
# numHHUS = [FRED Total Households (TTLHH)](https://fred.stlouisfed.org/series/TTLHH)
```

```{r getTotalHouseholdsFromFRED, echo=echoWorking}
# install.packages("fredr")
# update.packages("rlang") # latest version of rlang may also be necessary
library("fredr")
APIkeyPL = "96d09b567c2b44d33ffb53e34d47897a"
fredr_set_key(APIkeyPL) # Set the required FRED API key for the session.
# Can persistently set key by appending ""FRED_API_KEY=abcdefghijklmnopqrstuvwxyz123456 to an .Renviron file in the working directory or at the user level. 
# fredr() or fredr_series_observations() - Fetch a FRED series. By default, return a tibble
# fredr_series_search_text(search_text = "text sought") # - Search for a FRED series by text (in the series description). Returns tibble of matching series
# fredr_series_search_id(search_text = "IDsought", limit=NULL) - to search for a series by character ID of the series. The data (matching series) returned is a tibble
# fredr_series(series_id = "UNRATE") - function returns information for a single series specified by series_id. Returns tibble wit ro
# fredr_request() - Send a general request to the FRED API.


# Getting Started with fredr for FRED data: see https://cran.r-project.org/web/packages/fredr/vignettes/fredr.html

# TTLHH = fredr_series_search_text(
#   search_text = "Total Households",
#   order_by = "search_rank",
#   sort_order = "desc",
#   limit = 1) %>%
#   pull(id) %>%
#   fredr(series_id = .,
#         observation_start = as.Date("1990-01-01")
#   )
# 
# TTLHH %>%
#   ggplot(data = ., mapping = aes(x = date, y = value, color = series_id)) +
#     geom_line() +
#     labs(x = "Observation Date", y = "Number", color = "Series")

# head(TTLHH)
```

## Relationship Between the Elasticity of VKT and Elasticity of PKT Given Ride Pooling

Vehicle Kilometers Travel (VKT) or $M_V$ is related to Passenger Kilometers Travel (PKT) or $M_P$ through vehicle occupancy *o*

$$M_V = M_P /o$$
We take this as the defining relationship for average occupancy *o*. That is, *o* is the actual distance-weighted average of vehicle occupancy over all travel segments, including e.g. ZOV kilometers from repositioning, single-passenger kilometers and high-occupancy kilometers when pooling. As such is is the ratio of passenger km travel to vehicle km travel.^[A fine point is whether occupancy includes in the numerator PKT any passenger travel on added _detour_, or unproductive, km when deviating from their shortest O-D path to pick up or drop off other passengers. We argue it should not, and thus such detour miles lower the average effective occupancy.] Thus average occupancy can be less than or greater than the reference level for an average private household vehicle, $o_r$.

The average travel cost per passenger kilometer $C_P$ is vehicle (capital and operating) cost per kilometer $C_V$ shared over passengers plus

- time cost for each passenger along their intended route, given VoTT $C_T(o)$ [\$/hr] and average pace $\pi$ [hr/km]; 
- plus additional cost of passenger detour time in pooled mode, for incremental detour distance $\delta(o)$ [unitless];
- plus equivalent cost of the disutility of sharing space $C_D(o)$ as occupancy increases.

$$C_P(o) ~=~ C_V/o + C_T(o) \cdot \pi + C_T(o) \cdot \pi \cdot \delta(o) + C_D(o)$$
Posit for now the function $\gamma (o)$ which captures the fractional increase in cost-increase with occupancy, that includes the effect of those cost terms that are unrelated to vehicle capital and operating cost.

$$\gamma(o) ~=~ \frac {C_P(o)}{C_V/o} ~=~ 1 + \frac {\left ( C_T(o) \cdot \pi + C_T(o) \cdot \pi \cdot \delta(o) + C_D(o) \right )}{C_V/o} $$
It is not certain that total total cost will follow such a from in occupancy, but this is a reasonable approximation until more research/data on the effects of pooling on real and perceived costs is available.

So, on a per-passenger-kilometer cost basis

$$ C_P(o) ~=~ \frac {C_V} {o} \cdot \gamma (o)$$
where $\gamma (o_r) \equiv \gamma_r = o_r C_P/C_V$, the reference ratio of full passenger costs to vehicle capital/operating costs, and the first derivative $\gamma' (o) \ge 0$.

As written here, the vehicle costs $C_V$ refer _only_ to capital and operating costs per vehicle kilometer and exclude all time/disutility costs for the passengers. The vehicle related costs per passenger-km decline inversely with *o* while the time/disutility cost per passenger-km rises according to some monotonic function of *o*.

Q: Is the elasticity of VKT with respect to this measure of cost per vehicle kilometer useful?

Consider travel demand elasticities with respect to (generalized) travel cost

$$ElasPKT \equiv \epsilon_{M_P,C_P} \equiv \frac {d ln PKT} {d ln C_P}$$
is the elasticity of PKT w.r.t. passenger cost/mi
and 

$$ElasVKT \equiv \epsilon_{M_V,C_V} \equiv \frac {d ln VKT} {d ln C_V}$$
is the elasticity of VKT w.r.t. vehicle cost/mi.

#### What is the relationship between elasticities?

Recall that the elasticity of a function composition is the chained product of the elasticities of each function in the composition:

$$\epsilon_{f(y(x)),x} = \frac {d ln f(y(x))} {d ln x} = \frac { {df/f(y)} }{dx/x}  $$


$$\begin {align}
\epsilon_{f(y(x)),x} &= \frac {df}{dy}\frac {dy}{f(y)} \frac {x}{dx} \\
&= \frac {df}{dy}\frac {dy}{f(y)} \frac {y}{y} \frac {x}{dx} \\
&= \frac {df}{dy}\frac {y}{f(y)} \frac {dy}{dx} \frac {x}{y} = \epsilon_{f,y} \epsilon_{y,x}  
\end {align}$$


The demand for VKT is a function of the demand for PKT which is in turn a function of the cost of vehicle travel and occupancy

$$M_V(C_V) = f(M_P(C_P(C_V, o)))$$
specifically

$$M_V(C_V) = \frac {M_P(C_P(C_V, o))}{o(\cdot)}$$

$$\epsilon_{M_V, C_V} \equiv \frac {d ln M_V} {d ln C_V} ~=~ \frac {d ln M_P} {d ln C_V} - \frac {d ln o} {d ln C_V} ~=~ \epsilon_{M_P, C_V} - \epsilon_{o, C_V}$$
Focusing on the first term on the right-hand-side
$$\frac {d ln M_P} {d ln C_V} ~=~ \frac {d ln M_P} {d ln C_P} \cdot \frac {d ln C_P} {d ln C_V}$$

$$\epsilon_{M_P, C_V} ~=~ \epsilon_{M_P, C_P} \cdot \epsilon_{C_P, C_V}$$
Finally, looking at the elasticity of passenger travel cost with respect to vehicle travel cost:
$$C_P ~=~ \frac {C_V}{o(\cdot)} \gamma(o(\cdot))$$
Using the rules for the elasticity of function products and compositions:

$$\epsilon_{C_P, C_V} ~=~ 1 - \epsilon_{o, C_V} + \epsilon_{\gamma, o} \cdot \epsilon_{o, C_V}$$

$$\epsilon_{C_P, C_V} ~=~ 1 - \epsilon_{o, C_V} (1 - \epsilon_{\gamma, o})$$
This result has the following corner cases:
$\epsilon_{C_P, C_V} ~=~ 1$, i.e. $C_P/C_V = constant$ if

- if $\epsilon_{o, C_V} = 0$, that is if occupancy is invariant w.r.t. vehicle travel cost; then $C_P$ and $C_V$ move in direct proportion ($\epsilon_{C_P, C_V} = 1$) and Elas_VMT with respect to vehicle cost equals Elas_PMT with respect to passenger cost ($\epsilon_{M_V,C_V} = \epsilon_{M_P, C_P}$).
- if $\epsilon_{\gamma, o} = 1$, that is if $\gamma(o) = k o^1$ and the disutility multiplier of passenger costs rises proportionally with occupancy *o*, essentially offsetting the monetary cost reduction from spreading the vehicle costs over more passengers.

Returning to the relationship between `ElasVKT` and `ElasPKT`:
$$\begin {align}
\epsilon_{M_V,C_V} &= \epsilon_{M_P, C_V} - \epsilon_{o,C_V}\\
&= \epsilon_{M_P, C_P} \epsilon_{C_P, C_V} - \epsilon_{o,C_V}\\
\end {align}$$

$$\epsilon_{M_V,C_V} = \epsilon_{M_P, C_P} \cdot \left (1 - \epsilon_{o, C_V} \cdot (1- \epsilon_{\gamma,o}) \right ) - \epsilon_{o,C_V}$$

This relates the elasticity of VKT with respect to vehicle costs to the elasticity of PKT with respect to costs per passenger mile, with terms (elasticities) that relate to the response of occupancy $o$ to vehicle cost, and the response of proporational travel disutility cost $\gamma$ to occupancy.

Q: Is the elasticity $\epsilon_{o, C_V}$ well-defined? Or do we need to speak of some other relation like $\epsilon_{o, C_P}$, the elasticity of occupancy to passenger cost per km?

#### Numerical Evaluations

Test formulation assumes:

$$\epsilon_{M_P, C_P} = -1.0$$
Proportional added cost function $\gamma(o)$ is constant-elastic in occupancy *o*:
$$\gamma(o) = \gamma_r \cdot (o/o_r)^{\epsilon_{\gamma,o}}$$

Consider a rang of test values for occupancy *o* and the elasticity of cost increase to passengers:

*o* = 0.5 : 2.0 : 0.1

$\epsilon_{\gamma,o}$ = 0 : 2 : 0.1

The elasticity of occupancy with respect to vehicle cost is the result of the passengers' utility maximization problem, assuming that the ride is priced to the passenger in terms of the vehicle (capital and operating) cost divided by occupancy, and the passengers take into account all the time and disutility cost associated with pooling.
$$o(C_V) = ???$$

```{r calcElasVPTfromPKT, echo=echoWorking}
elas_Mp_Cp = -1.0
o_vals = seq(from = 0.5, to = 2.0, by = 0.1)  # not used in elasticity calculation
elas_gamma_o = seq(from = 0, to = 2.0, by = 0.1)

elas_o_Cv_vals = seq(from = 0, to = 0.5, by = 0.1)

elas_Mv_Cv = matrix(data=NA, nrow = length(elas_gamma_o), ncol= length(elas_o_Cv_vals))
# data_frame(elas_gamma_o = seq(from = 0, to = 2.0, by = 0.1))

for (i in seq_along(elas_o_Cv_vals))  {
  elas_Mv_Cv[,i] = elas_Mp_Cp * (1 - elas_o_Cv_vals[i] * (1- elas_gamma_o) ) - elas_o_Cv_vals[i]
}

elas_Mv_Cv = as.data.frame(x = elas_Mv_Cv)
#names(elas_Mv_Cv) = paste("elas_o_Cv", elas_o_Cv_vals, sep="#")
names(elas_Mv_Cv) = elas_o_Cv_vals
elas_Mv_Cv$elas_gamma_o = elas_gamma_o

elas_Mv_Cv = elas_Mv_Cv %>% 
  gather(key = elas_o_Cv, value = elas_Mv_Cv, -elas_gamma_o) # gather up all columns except elas_gamma_o
# tail(elas_Mv_Cv)

elas_Mv_Cv %>% ggplot(aes(x=elas_gamma_o, y= elas_Mv_Cv, color=elas_o_Cv)) + 
  geom_line() + ggtitle("Variation in Elasticity-of-VMT/Elasticity-of-PMT with elas_gamma_o")
```

## 2. PMT and VMT for Various Levels of Pooling - Numerical Examples

```{r setDemandAndPoolingParams, echo=echoWorking}
ElasPKT	= -1	            #	e_{PMT,CT}
ElasCTOccupancy	= 0.5	    #	elas_{C_T,o}
ReductCTAutomation	= 0.5	#	rho_A
ReductPaceAutomation	= 1	#	PaceReduct_A
ElasTourContractionWithPooling = 0.0 #
ElasRiderDetourWithPooling = 0.0
PacePi =1/39.98
```

### Read cost informtion (vehicle travel cost per mile by category) for Base assumptions and Alternative assumptions.

```{r readBaseAndAltCostTables, echo=echoWorking}
VTCostBase = read.csv("VTCostBase.csv", comment.char = "#")
VTCostAlt = read.csv("VTCostAlt.csv", comment.char = "#")

VTCost = VTCostAlt %>% select(DemScen, VehType, CostCat, VTCost_cpm) # extract essential vars
VTCostBase$DemScen="Base"              # label base as another scenario
VTCostBase = VTCostBase[names(VTCost)] # order cost to match alt cost scenario data
VTCost = rbind(VTCost, VTCostBase)

VTCost1 = VTCost %>% dplyr::filter(
  VehType == "LDVAvgSedan"
  ) %>%
  spread(key = CostCat, value = VTCost_cpm)

```

```{r calcPMTandVMT0, echo=echoWorking}
PMTVMT0 = data.frame("o" = seq(from = 0.8, to = 1.5,by = 0.05)) #	occupancy, unitless (pass/veh, normalized)
#M = M_b/o		#	unitless (normalized VMT)
#M_B /o =1/o		#	unitless (normalized VMT)

VTCost1 = VTCost1 %>%
  rename(C_F = Fuel,         # fuel costs, c/mi
         C_T = Time) %>%     # time costs, c/mi
  mutate(
    C_H = Total - C_F - C_T, # non-fuel vehicle financial costs
    C_V = C_F + C_H		       # vehicle costs, c/mi
  )

PMTVMT0 = PMTVMT0 %>% mutate(
  C_Fbase = VTCostBase1$Fuel,		#	cents/mi VTCostBase(Fuel, LDVAvgSedan)
  C_F = VTCostAlt1$Fuel,		#	cents/mi VTCostAlt(Fuel, LDVAvgSedan)
  C_Hbase = VTCostBase1$Total - VTCostBase1$Fuel - VTCostBase1$Time,		#	cents/mi VTCostBase(NonFuelOrTime, LDVAvgSedan)
  C_H = VTCostAlt1$Total - VTCostAlt1$Fuel - VTCostAlt1$Time,  		#	cents/mi VTCostAlt(NonFuelOrTime, LDVAvgSedan)
  C_Tbase = VTCostBase1$Time	  
)

PMTVMT0 = PMTVMT0 %>% mutate(
  C_Vbase = C_Fbase + C_Hbase,		#	cents/mi
  C_V = C_F + C_H,		#	c/mi
  c_Vbase = C_Vbase / C_Vbase[o==1.0],		#	unitless
  c_V = C_V /C_Vbase[o==1.0] 	#	unitless
)

PMTVMT0 = PMTVMT0 %>% mutate(
  C_VPM_base = C_Vbase /o,		#	c/pass-mi
  C_VPM = C_V / o,		#	c/pass-mi
  rho_o = o^ElasCTOccupancy,		#	unitless (Pooling Time Cost Mult)
  Pi_base =1/39.98,		#	hr/mi
  Pi = Pi_base * ReductPaceAutomation		#	hr/mi
)

PMTVMT0 = PMTVMT0 %>% mutate(
  C_Tbase = rho_o *C_Tbase/Pi_base,		#	cents/hr
  C_T = C_Tbase*ReductCTAutomation		#	cents/hr
  #"= C_Tbase*Pi_base,		#	cents/mi
  #"= C_T * Pi,		#	cents/mi
)

PMTVMT0 = PMTVMT0 %>% mutate(
  C_Mbase = C_Vbase/o + C_Tbase*Pi_base,		#	cents/mi
  C_M = C_V/o + C_T*Pi		#	cents/mi
)

PMTVMT0 = PMTVMT0 %>% mutate(
  PMT_mv = (C_Mbase/C_Mbase[o==1.0])^ElasPKT,		#	pass-mi (normalized)
  PMT_av = (C_M/C_Mbase[o==1.0])^ElasPKT,		#	pass-mi (normalized)
  VMT_mv = PMT_mv/o,		#	veh-mi (normalized)
  VMT_av = PMT_av/o,		#	veh-mi (normalized)
  VMT_mv_direct = PMT_mv[o==1.0]/o,		#	veh-mi (normalized)
  VMT_av_direct = PMT_av[o==1.0]/o		#	veh-mi (normalized)
)

```

```{r setDemandAndPoolingParams1, echo=echoWorking}
ElasPKT	= -1	#	e_{PMT,CT}
ElasCTOccupancy	= 0.5	#	elas_{C_T,o}
ReductCTAutomation	= 0.5	#	rho_A
ReductPaceAutomation	= 1	#	PaceReduct_A
ElasTourContractionWithPooling = 0.0 #
ElasRiderDetourWithPooling = 0.0
PacePi =1/39.98
```


```{r calcPMTandVMT, echo=echoWorking}
VTCostBase = read.csv("VTCostBase.csv", comment.char = "#")
VTCostAlt = read.csv("VTCostAlt.csv", comment.char = "#")

VTCostAlt1 = VTCostAlt %>% filter(
  DemScen == "DS3",
  VehType == "LDVAvgSedan"
  ) %>%
  select(CostCat, VTCost_cpm) %>%
  spread(key = CostCat, value = VTCost_cpm)
VTCostBase1 = VTCostBase %>% filter(
  VehType == "LDVAvgSedan"
  ) %>%
  select(CostCat, VTCost_cpm) %>%
  spread(key = CostCat, value = VTCost_cpm)


PMTVMT = data.frame("o" = seq(from = 0.8, to = 1.5,by = 0.05), Case0="base", Case1="alt") #	occupancy, unitless (pass/veh, normalized)
#M = M_b/o		#	unitless (normalized VMT)
#M_B /o =1/o		#	unitless (normalized VMT)


PMTVMT = PMTVMT %>% mutate(
  C_Fbase = VTCostBase1$Fuel,		#	cents/mi VTCostBase(Fuel, LDVAvgSedan)
  C_F = VTCostAlt1$Fuel,		#	cents/mi VTCostAlt(Fuel, LDVAvgSedan)
  C_Hbase = VTCostBase1$Total - VTCostBase1$Fuel - VTCostBase1$Time,		#	cents/mi VTCostBase(NonFuelOrTime, LDVAvgSedan)
  C_H = VTCostAlt1$Total - VTCostAlt1$Fuel - VTCostAlt1$Time,  		#	cents/mi VTCostAlt(NonFuelOrTime, LDVAvgSedan)
  C_Tbase = VTCostBase1$Time	  
)

PMTVMT = PMTVMT %>% mutate(
  C_Vbase = C_Fbase + C_Hbase,		#	cents/mi
  C_V = C_F + C_H,		#	c/mi
  c_Vbase = C_Vbase / C_Vbase[o==1.0],		#	unitless
  c_V = C_V /C_Vbase[o==1.0] 	#	unitless
)

PMTVMT = PMTVMT %>% mutate(
  C_VPM_base = C_Vbase /o,		#	c/pass-mi
  C_VPM = C_V / o,		#	c/pass-mi
  rho_o = o^ElasCTOccupancy,		#	unitless (Pooling Time Cost Mult)
  Pi_base =1/39.98,		#	hr/mi
  Pi = Pi_base * ReductPaceAutomation,		#	hr/mi
  TourContraction = o^ElasTourContractionWithPooling,
  RideDetourExpansion = o^ElasRiderDetourWithPooling
)

PMTVMT = PMTVMT %>% mutate(
  C_Tbase = rho_o *C_Tbase/Pi_base,		#	cents/hr
  C_T = C_Tbase*ReductCTAutomation,		#	cents/hr
  #"= C_Tbase*Pi_base,		#	cents/mi
  #"= C_T * Pi,		#	cents/mi
  C_Mbase = C_Vbase/o + C_Tbase*Pi_base,		#	cents/mi
  C_M = C_V/o + C_T*Pi		#	cents/mi
)

PMTVMT = PMTVMT %>% mutate(
  PMT_mv = (C_Mbase/C_Mbase[o==1.0])^ElasPKT,		#	pass-mi (normalized)
  PMT_av = (C_M/C_Mbase[o==1.0])^ElasPKT,		#	pass-mi (normalized)
  VMT_mv = PMT_mv/o,		#	veh-mi (normalized)
  VMT_av = PMT_av/o,		#	veh-mi (normalized)
  VMT_mv_direct = PMT_mv[o==1.0]/o,		#	veh-mi (normalized)
  VMT_av_direct = PMT_av[o==1.0]/o,		#	veh-mi (normalized)
  EnergyUse_av = VMT_av*C_F/C_Fbase
)
```

```{r test1}
More = PMTVMT %>% select(o, C_Vbase, C_Tbase, C_V, C_T, Pi_base, Pi)
vals = PMTVMT %>% select(o)
vals$ElasTourContractionWithPooling = seq(from = -0.3,to = 0,length.out = nrow(vals))

left_join(More, vals, by="o")

```

