---
title: "Notes on Valuation of Travel Time"
author: "Paul Leiby"
date: "Mar 29, 2017"
output:
  word_document:
    fig_caption: yes
    fig_height: 5
    fig_width: 7
  pdf_document: default
---

Paul N. Leiby
Oak Ridge National Laboratory
Leibypn@ornl.gov
865.574.7720 


```{r setChunkOptions, echo=FALSE}
## clear: remove (almost) everything in the working environment.
## You will get no warning, so don't do this unless you are really sure.
rm(list = ls())

# Note: knitr options are:
#  opts_knit$set(root.dir = '../..')
# include = FALSE prevents code and results from appearing in the finished file.
#   R Markdown still runs the code in the chunk, and the results can be used by other chunks.
# echo = FALSE prevents code, but not the results from appearing.
#   This is a useful way to embed figures.
# message = FALSE prevents messages generated by code from appearing.
# warning = FALSE prevents warnings generated by code from appearing
# fig.cap = "..." adds a caption to graphical results.

# Logical Flags to indicate which chunks should be displayed in RMarkdown output
echoWorking = FALSE # set this level to FALSE for an exported (non-working version) of output
echoMaximal = FALSE # set TRUE only if you want echo just about everything
```

```{r setWorkingEnv, echo=echoWorking}
# opts_chunk$set(comment=NA, fig.width=6, fig.height=6)
# opts_chunk$set(dpi=150)

#machine = "PLWinLaptop"
#machine = "PLWinBox"
machine = "PLMac"
worksubpath = "/Papers/2015SMARTInitiative/2017SmartMobilityMDS/Analysis/"
datasubpath = paste0(worksubpath,"Data/")  # subdirectory for data

if (machine == "PLWinBox") {
  workroot = "D:/"  # for Windows OS, Paul's work desktop
} else if (machine == "PLRemoteSiegfried") {
  workroot = "C:/"  # for Windows OS, remote into Siegfried
} else if (machine == "PLWinLaptop") {
  workroot = "C:/" # for Windows OS on laptop
} else if (machine == "PLMac") {
  workroot = "~/Documents/"  # for OSx Paul's Mac  
} else if (machine == "JRLaptop") {
  workroot = "C:/"  # JR - set this to any machine-specific prefix to the worksubpath above 
}
```

Notes on Value of Travel Time and Vehicle Automation
=================================

1. Existing estimates of Travel Time Value
2. Theoretical Basis of Travel Time Value
    1. dimensions and determinants of travel time value
3. Empirical Methods of Estimating Travel Time Value 
4. Data Sources for Estimating Travel Time Value
5. Applications and Extensions of Value of Travel-Time Theory to Travel in CAVs
6. Ongoing/Initial VOTT Research for CAVs

----------------------

We review the conceptual issues associated with Value of Travel Time (VOTT), from the existing literature. Established estimates of VOTT and guidance for their application are offered periodically by the U.S. DOT (e.g. White/DOT 2016, Endorf/DOT 2015).


1. Existing Estimates of Travel Time Value
-----------------
The value of travel time (VOTT) is a major determinant of existing transportation system. A key conclusion is that despite extensive research and progress, the concepts of VOTT are in flux and challenged, both theoretically and empirically. Small, in his 2012 review, concludes: "value of travel time remains incompletely understood and ripe for further theoretical and empirical investigation."
A second conclusion from the literature is that VOTT is heterogenous, varying across individual, travel type, and travel conditions. "A recurrent theme is that the value of time for a particular travel movement depends strongly on very specific factors, and that understanding how these factors work will provide new insights into travel behavior and into more general economic choices." [Small 2012:2] The U.S. DOT, for example makes important distinctions in VOTT on basis of

* Activity/trip purpose (business or personal)
* Driving Conditions (local or intercity, congested or free flow)
* Traveler type (car, bus, truck driver, pedestrian, transit rider)
        
The established literature recognizes the need to differentiate between Value of Travel Time VOTT and Value of Reliability VOR. The latter is a related but separable concept from the former. This distinction is particularly relevant in the context of CAVs, because CAVs can alter both the mean duration and quality of travel time, as well as increase the reliability of travel time.

### Issues and Challenges in Analyzing Changes to the Valuation of Travel Time, and Their Implications
* How can we best organize the dimensions of VOTT, for the purposes of its estimation in the context of CAVs?
* Given changing estimates of VOTT, what special considerations, if any, apply to the use of VOTT measures in modeling CAV travel demand response?
* What are the approaches for the effective use of VOTT in modeling and estimating CAV travel demand response?
* Do the theoretical foundations of the VOTT concept, as used in travel demand models, need revisiting for CAVs? CAVs, in combination with other information technologies, have the potential to alter not ony travel behavior, but also how people decide on their non-travel activities, and schedule them. We need to be attentive to potential changes that may be needed in modeling foundations, such as the treatment of travel as a derived demand, and travel time as one of the competing uses for time in a finite-time budget allocation problem.
    * If CAVs significantly alter planning and travel behavior, then implications for travel demand may not be inferable from CAV-induced changes in generalized travel cost, even when including the revised valuation of travel time.


### Observations on Travel Time Value Estimates

Small (2012) offers a number of observations on existing travel time estimates. The value of time for commute trips are found to typically average around one-half the gross wage. Those values also vary widely by trip purpose. Typically values are highest for business travel and lowest for discretionary or leisure travel. This is reflected in values recommended in the U.S. DOT guidance on travel time valuation (White/DOT 2016, Ayala/DOT 2014). Estimate values in the literature rise with income or wage rate, but less than proportionally (i.e., income elasticity of travel time value is less than 1). Income elasticities have been reported between 0.5 and 0.9, based on time-series and cross-sectional studies. 

Variations in travel time valuation across different travel types and travel conditions are well-established. VOTT is higher under congested conditions than free-flow conditions, by ~25% to 55%. For transit trips, walking or waiting time is valued much higher than the time spent in the vehicle (by factor of 2- 3). Studies also document additional variation in VOTT, called "unobserved heterogeneity," which is not explainable by observed travel attributes.  For this reason, the currently preferred estimation approach for VOTT is to use a mixed discrete choice model. These models are a generalization of the standard discrete-choice models (such as probit, logit, or generalized extreme value) that invoke individual maximization of random utility. For mixed discrete models the error term has 2 random components, one associated with a standard discrete-choice model, a second general component that depends on the data. The second random distribution effectively reflects uncertainty or randomness in the coefficients for explanatory factors, allowing random taste variation and unrestricted subtitution patterns (Train 2003).
    
### VOTT Estimates from U.S. DOT

U.S. DOT guidange on the valuation of travel time offers these principles (Ayala 2014):

> "The value of reducing travel time expresses three principles. First, time saved from travel could be dedicated to production, yielding a monetary benefit to either travelers or their employers. Second, it could be spent in recreation or other enjoyable or necessary leisure activities for which individuals are willing to pay. Third, the conditions of travel during part or all of a trip may be unpleasant and involve tension, fatigue, or discomfort.""

Thus the  travel time value reflects the _opportunity cost_ of time, _plus_ the disutility (discomfort) associated with travel. This is consistent with the Becker utility-based model and its decendants, described below. The 2016 DOT recommended values and ranges are show in the figure below.
```{r DOT_VOTT_ests, echo=echoWorking}
library(knitr)  # for kable

# Source: US DOT: “Plausible Ranges for Hourly Values of Travel Time Savings” (2009 U.S. $ per person-hour)
#  see "Value Of Travel Time.xlsx"
# VOTT_DOT <- read.delim(textConnection("
# Type	     Purpose	      Type_Purpose        High	  Low	    Recommended
# Local	     Personal      Local_Personal      14.30 	8.40 	12.00 
# Local	     Business      Local_Business      27.50 	18.30 	22.90 
# Local	     AllPurposes 	Local_AllPurposes 	14.90 	8.90 	12.50 
# Intercity  Personal	    Intercity_Personal	21.50 	14.30 	16.70 
# Intercity  Business	    Intercity_Business	27.50 	18.30 	22.90 
# Intercity  AllPurposes 	Intercity_AllPurposes 	22.80 	15.20 	18.00 
# NA         TruckDrivers	TruckDrivers	29.60 	19.80 	24.70 
# NA         BusDrivers	  BusDrivers	29.40 	19.60 	24.50 
# NA         Walk_access_waiting_transfer Walk_access_waiting_transfer	28.70 	19.10 	23.90
# "),header=TRUE,sep="",strip.white=TRUE)
# Source: US DOT 2016: “Plausible Ranges for Hourly Values of Travel Time Savings” (2015 U.S. $ per person-hour)
VOTT_DOT <- read.delim(textConnection("
Type	     Purpose	      Type_Purpose          High	  Low	    Recommended
Local	     Personal      Local_Personal         16.30 	 9.50 	13.60 
Local	     Business      Local_Business         30.50 	20.30 	25.40 
Local	     AllPurposes 	Local_AllPurposes 	    17.00 	10.00 	14.10 
Intercity  Personal	    Intercity_Personal	    24.50 	16.30 	19.00 
Intercity  Business	    Intercity_Business	    30.50 	20.30 	25.40 
Intercity  AllPurposes 	Intercity_AllPurposes 	25.80 	17.20 	20.40 
Other      TruckDrivers	TruckDrivers	          32.70 	21.80 	27.20 
Other      BusDrivers	  BusDrivers	            34.00 	22.70 	28.30 
Other      Walk_wait    Walk_wait	              32.60 	21.70 	27.15
"),header=TRUE,sep="",strip.white=TRUE)

kable(VOTT_DOT, caption = "Table: DOT Estimates of VOTT")

```

```{r plotVOTTranges, DPI=300}
library(tidyverse)
myData = VOTT_DOT %>%
  dplyr::select(Type_Purpose, High, Low, Recommended)
  # gather(value=estimates, key= High:Recommended) 
myData2 = VOTT_DOT %>%
  dplyr::select(Type, Purpose, High, Low, Recommended)

# geom_boxplot(lwd=3) ("lwd" for "line width"). 
# Also, if lwd makes the median too thick, you can use lwd and fatten
# together to make the median line thinner relative to the other lines

# ggplot(data=myData, aes(x=Type_Purpose, 
#                         ymin = `Low`, 
#          ymax = `High`, lower = `Low`, 
#          upper = `High`, middle = `Recommended`)) + 
#      geom_boxplot(stat = 'identity',  
#                   fill = "lightblue", colour="lightblue",
#                   position = position_dodge(1), width=0.5) +
#      geom_crossbar(aes(y = `Recommended`), colour="black", fatten=2) +
#      xlab("") +
#      ylab('$/hour') +
#      ggtitle("DOT VOTT Estimates: Recommended Values with Ranges")

# Grouped, 2-factor bar plot with ranges
# See http://stats.stackexchange.com/questions/11406/boxplot-with-respect-to-two-factors-using-ggplot2-in-r
ggplot(data=myData2, aes(x=Type,
                         y=Recommended,
                        fill = Purpose,
                        ymin = `Low`, 
         ymax = `High`, lower = `Low`, 
         upper = `High`, middle = `Recommended`)) + 
     ylab('$/hour') +
     geom_boxplot(stat = 'identity') +
                  # position = position_dodge(1), width=0.5) +
     # geom_crossbar(aes(y = `Recommended`), colour="black", fatten=2) +
     ggtitle("DOT VOTT Estimates: Recommended Values with Ranges")

```


2. Theoretical Basis of Valuation of Travel Time
-------------------------

### Definition of the value of travel time 
The value of travel time (VTT, or $V_T$) is defined as the "tradeoff between time and cost leaving utility constant" (Lave 1969, Small 2012), i.e. the marginal rate of substitution between time and monetary cost.  This may be computed as or marginal (conditional, indirect) utility of travel time divided by the marginal utility of income (or cost/dollars)
$$V_T \equiv \frac{(\partial V/\partial T_T)}{(\partial V/ \partial C)}$$
For example, if conditional indirect utility for a transportation mode is linear in cost and travel time (as in Lave 1969 model):
$V = \alpha C + \beta w T_T$
where *C* is cost, $T_T$ is travel time, and *w* is wage rate, 
then VTT is $(\partial V/\partial T_T)/(\partial V/ \partial C) = (\beta/\alpha)w$.

We note that one important effect of CAV technology is to change the marginal utility of travel time for any level of travel, presumably increasing the direct consumption utility (the inherent enjoyment of travel) and reducing the opportunity cost.

### VOTT Theory Based on the Traveler's Utility Maximization Problem
The classic theoretical basis for VOTT is based on the utility maximization problem of the traveler. The standard Becker (1965) model of time allocation across work, leisure and travel, as elaborated by DeSerpa (1971), accounts for multiple non-work activities $i=1..K$ including $T_T$ being the time allocated to travel and $T_w$ the time allocated to work. The income from labor plus unearned income *Y* is used to purchase goods *G*.  There are constraints on budget (where $P_G \equiv 1$), total time expenditure, and constraints indicating the minimum allowable time to be spent on each activity: 
$$
\begin{aligned}
Max_{G,T_1, ...T_K,T_w} & U(G,T_1, ...T_K,T_w )  &\\
s.t. ~ & G \le w * T_w + Y & : \lambda_Y\\
& T_w + \Sigma_{i=1}^K T_i \le \bar T & : \lambda_T \\
& T_i \ge \bar T_i ~\forall i \in (1..K) & : \phi_i\\
\end{aligned}
 $$
 With the Lagrangian:
$$
\begin{aligned}
L = & U(G,T_1, ...T_K,T_w ) \\
&+ (G - w * T_w - Y) \cdot \lambda_Y\\
&+ (T_w + \Sigma_{i=1}^K T_i - \bar T) \cdot \lambda_T \\
&+ \Sigma_{i=1}^K (\bar T_i - T_i) \cdot \phi_i\\
\end{aligned}
$$
The optimality conditions, at maximizing point $(G^*,T_1^*, ...T_K^*,T_w^*)$ are:
$$\begin{aligned}
\frac {\partial L}{\partial G} &= \frac {\partial U}{\partial G} + \lambda_Y = 0\\
\frac {\partial L}{\partial T_w} &= \frac {\partial U}{\partial T_w} - w \lambda_Y + \lambda_T + \phi_w = 0\\
\frac {\partial L}{\partial T_i} &= \frac {\partial U}{\partial T_i} + \lambda_T - \phi_i = 0 ~~ \forall i \in (1..K)\\
\end{aligned}
$$

For $V(Y, \bar T, \bar T_T, w) \equiv U(G^*,T_1^*, ...T_K^*,T_w^* | Y, \bar T, \bar T_i, w)$ being the maximal (indirect) Utility for given constraint levels and wage $w$, the sensitivity theorem says that the Lagrange multipliers provide the marginal variation in indirect utility with constraint RHS values:
$$\begin{aligned}
\frac {\partial V}{\partial Y} &= -\lambda_Y \ge 0 \\
\frac {\partial V}{\partial \bar T} &= -\lambda_T \ge 0 \\
\frac {\partial V}{\partial \bar T_i} &= -\phi_i \ge 0 ~~ \forall i ~???\\
\end{aligned}
$$
Here the Lagrange variables are all defined with the following signs, 
$$\begin{aligned}
\lambda_Y \le 0 \\
\lambda_T \le 0 \\
\phi_i \le 0 ~~ \forall i ~???\\
\end{aligned}
$$

Combining these relationships we can find an expression for the marginal (indirect) utility of travel time at the optimum (this exposition follows Small and Verhoef 2007 Chapter 2):
$$
\begin{aligned}
-\frac{\partial V}{\partial \bar T_T} = \phi_T  &= - \frac{\partial U}{\partial T_T} - \lambda_T \\
&= -\frac{\partial U}{\partial T_T} - \left ( - \frac{\partial U}{\partial T_w} + w \lambda_Y  \right )
\end{aligned}
$$
(the second expression assuming lower bound on work is non-binding and $\phi_w=0$, or equivalently, work time is not in the set of lower-bounded activities)
and
$$\phi_T = -\lambda_T - \frac{\partial U}{\partial T_T} \equiv -\lambda_T - U_{TT}$$
$$ -\lambda_T = U_{TW} - w \lambda_Y$$
this condition considers that direct utility of labor as well as its wage, balanced with the shadow cost of the time budget constraint.
$$\phi_T = U_{Tw} - w \lambda_Y - U_{TT} $$

which leads to an expression for the Value of Travel Time VTT
$$V_{TT} \equiv \frac{(\partial V/\partial T_T)}{(\partial V/ \partial C)}$$

$$ 
V_{TT} = \frac {\phi_T}{-\lambda_Y} = w + \frac {U_{Tw} - U_{TT}}{-\lambda_Y}
$$
(See Small 2012:4).

Thus, for this model where travel time displaces work time and other activities, the value of travel time is equal to the wage, adjusted by a term that accounts for the difference between the marginal utility (direct consumption benefit) of work time and the marginal utility of travel time (with the marginal utility difference converted to dollars by the marginal utility of (unearned) income, $-\lambda_Y$). 

Note that this expression accounts for both the opportunity cost of travel time in terms of foregone wage ($w$) and the relative consumption benefit of work time versus travel time.

For this model a similar condition would apply to the value of time spent on all other non-work activities. E.g., for activity *i*
$$ 
V_{Ti} = \frac {\phi_T}{-\lambda_Y} = w + \frac {U_{Tw} - U_{Ti}}{-\lambda_Y}
$$


Q: What about taxation on labor? That alters the optimal labor level and would pre-multiply wage rate $w$ in the value of travel time. Wage $w$ here should be viewed as after-tax wage.
Q: What about the utility derived from the trip itself? This formulation conflates the direct consumption benefit of time spent traveling itself with the benefit of what travel time produces (e.g. VMT and arriving at the trip destination). The model in Parry and Small (2015) offers one representation that separates those attributes, and their utilities.

In general, most forms of travel time are evidently more pleasant than work time, because the estimated value of travel time is typically below the wage indicating that the second adjustment term is negative.

Or (alt derivation)???
$$
\begin{aligned}
V_T \equiv -\frac {\partial V / \partial \bar T_T} {\partial V /\partial Y}
&= \frac {\phi_T}{\lambda_Y}  \\
&= \frac {\frac{\partial U}{\partial T_T} - w \frac{\partial U}{\partial G}}{\frac{\partial U}{\partial G}}  ???
\end{aligned}
$$

### Other Theoretical Variations


Henscher’s (1977) model of business travel accounts for multiple value-attributes of business travel time,in a way that has some analog to the anticipated multi-purpose travel time in automated vehicles. He discussed the value of time saving to be considered by employers as employees trade off to work vs. travel vs. working while travelling. The model accounts for productive time while traveling (likely at reduced productivity), as well as the reduced productivity of work time following extended travel and fatigue. This latter factor of travel fatigue could also serve as a limit on the desired extent of travel in CAVs.

Small (2012:5) notes that Henscher's model business travel time and models of travel time types have generally been difficult to estimate empirically, due the role of factors that are difficult to observe:
> The value of time that affects travel decisions can depend strongly on factors that analysts normally do not observe: individual tax rates, ability to use travel time productively, fatigue from travel, enjoyment of work, tightness of work-hours constraints, and many others. 

Small (2012): "For reviews of additional theoretical developments, see Gonzalez (1997), Jara-Diaz (2000, 2007), and Hensher (2011)."

* Gonzalez, Rosa Marina, 1997. The value of time: a theoretical review. Transport Reviews 17 (3), 245–266.
* Hensher, David A., 2011. Valuation of travel time savings. In: de Palma, A., Lindsey, R., Quinet, E., Vickerman, R. (Eds.), A Handbook of Transport Economics.
Edward Elgar, pp. 135–159. (Chapter 7).
* Jara-Diaz, Sergio, 2000. Allocation and valuation of travel-time savings. In: Hensher, D.A., Button, K.J. (Eds.), Handbook of Transport Modeling. Pergamon, Amsterdam, pp. 303–319.
* Jara-Diaz, Sergio, 2007. Transport Economic Theory. Elsevier,Amsterdam. 
* Jara-Diaz, S.R., Munizaga, M.A., Greeven, P., Guerra, R., Axhausen, K., 2008. Estimating the value of leisure from a time allocation model. Transportation
Research Part B 42 (10), 946–957.

Small (2012): "Jara-Diaz et al. (2008) provide more direct evidence for this interpretation, as they develop a method to separately estimate the value of travel time $V_{TK}$ and the value of leisure $\mu/\lambda$."

VOTT and VMT Implications of Congestion
----------------------

CAVs are expected to both make traveling in congested traffic much more pleasant (reducing both stress and boredom), while CAVs collectively are expected to reduce the levels of congestion through traffic coordination and dynamic re-routing. Travel time in congested travel is typically estimated to have much higher costs than in free-flow, so the implications of CAVs for VOTT are inextricably linked with their implications for congested travel. For example, Fagnant and Kockelman (2015) estimate the induced VMT effects of automated vehicles, accounting for two factors: additional unoccupied travel of shared AVs when traveling to pick up riders, and VMT induced by reduced costs and congestion. They estimate the former at approximately 10% for a mature/high-CAV-penetration market, based on simulations with their own agent-based model for the Austin, Texas market.  For the effects of congestion-reduction and increased highway effective capacity, they apply Cervero's (2001) survey and meta-analysis results for the urban area elasticity of VMT with respect to highway lane miles. For high (90 percent) CAV market penetration, they estimate 60% reduction in freeway congestion and 15% reduction in arterial congestion, leading to a 26% increase in VMT. These types of results provided estimate that could be compared with those based on changes in the quantity and valuation of congested travel time.

3. Empirical Methods of Estimating Travel Time Value 
------------------------
### Stated preference surveys relevant to CAV time use and valuation
Cyganski et al’s (2014) quantitative online survey in Germany, analyzed with ordered probit model, measured ordinal responses that allow inference of priorities and values, but not quantitative measures such as VOTT. Zmud et al. (2015, 2016) conducted online surveys followed by qualitative interviews to ascertain the impact of CAV technologies on travel behaviour.  This work is indicative, but would need to be extended to quantify either time use or time valuation.

The WholeTraveler Survey will XXX …

### Revealed Preferences and WTP Studies for Travel Time Value

WTP data for the estimation of VOTT can be derived based on driver choices given congestion pricing or toll lane pricing. For example, Sheikh et al. 2014 used toll data from Atlanta express lane corridor, considering travel time and reliability measures. Revealed preferences from mode choice can also suggest VOTT measures that may provide informative bounds for the VOTT in CAVs, given appropriate attention to the other attributes of the modes (e.g. privacy, comfort, reliability). Particularly informative is the choice between driving vs non-driving modes (taxi, ride service, metro, train). Ayala/DOT 2014 notes that VOTT or Value of Travel Time Savings (VTTS) research "is often based on the factors influencing mode choice, including the comfort, privacy, and prestige subjectively ascribed to particular modes, as well as travel time and cost."

### Combined Stated and Revealed Preference Studies
Small and Winston 2005, among others, have also effectively combined stated and revealed preference approaches in mixed logit study of travel responses to congestion pricing. 


### Mixed Discrete Choice Model

Current methods for estimating VOTT model discrete choice, commonly recognize "unobserved heterogeneity" by specifying that the known key explanatory variables (e.g. travel time, reliability, cost) have  *random* coefficients. These variables may be interacted with attributes specific to the traveler of travel type (income, congestion, purpose, distance, etc.) to account for observed heterogeneity.

Standard approach in VOTT studies is Mixed Discrete Choice Model, to reflect unexplained/unobserved sources of heterogeneity with randomness in the coefficients of the discrete choice model. For example, a "mixed logit model" 
$$V_{ij} = \alpha_j + \beta_0' z_{ij} + v_{ij}z_{ij} + \epsilon_{ij}$$
where e{ij} is a pure random term with an extreme value distribution (hence ‘‘logit’’ in the model’s name) and nij is a random function with some ‘‘mixing distribution,’’ so called because the resulting choice probabilities are integrals involving a mixture of the two distributions. Common distributions assumed for nij include normal, half-normal, log-normal, and triangular—the latter three often chosen in order to restrict coefficients to have their theoretically expected signs. 

Estimation is done by Monte Carlo simulation, simulating the log-likelihood function, which is maximized to estimate parameters $\beta_0$ and the "hyperparameters" of the mixing distribution. (see Train 2010, Small 2012:5-6). Various functional forms may be assume for the function $v_{ij}$, e.g. commonly
$$v_{ij} = \Sigma_k \sigma_k \phi_k z_{ij}^k$$
where v_{ij} is included to reflect the randomness in the coefficients $\beta_k$ of particular variables, $z^k$,

#### Challenges for Mixed Discrete Models
Mixed discrete models require the choice of a distribution function for the random coefficients. This is commonly done with non-parametric estimation (e.g. Fosgerau 2006) or "latent class model" (Greene and Hensher 2003)
* tendency to produce "implausible distributions of ratios," which is problematic since VTT is a ratio of estimated parameters (see discussion under theory)
    * denominator is typically the coefficient of a price/cost variable (corresponding to indirect marginal utility of income?)
    * address by making cost coefficient non-random?


4. Data Sources for Estimating Travel Time Value
------------------------
* Identifying data sources
	* Consider existing data sources that might be used/interpreted differently
	* Consider data on Proxy activities (taxi rides, Lyft, trains)

* Mode choice data to estimate revealed preferences
    * drive vs non-driving modes (taxi, ride service, metro, train)
    * distinctions between drivers and passengers
* Stated preference survey data
    * Small and Winston 2005 combined stated and revealed preference approaches in mixed logit study of responses to congestion pricing
* WTP Data based on driver choice given congestion or toll lanes pricing
    * Sheikh et al. 2014. Used toll data from Atlanta express lane corridor, considering travel time and reliability measures
* Trip-based data
* Data that influence VTT (observable sources of heterogeneity)
    * Type of vehicle (LDV, MDV, HDV)
    * Trip purpose (principal distinction made by DOT is business travel vs. personal (Ayala/DOT 2014) )
    * Trip duration/distance
    * Travel conditions
        * Activities while traveling?
    * Mode
    * Cost measures
    * Traveler demographic characteristics
        * DOT finds age, sex, education, employment "unlikely to practical" for VOTT application
        * distinctions between drivers and passengers thought more relevant
        * income
   
* Oregon DOT 2014. [The Value of Travel-Time: Estimates of the Hourly Value of Time for Vehicles in Oregon 2013](https://www.oregon.gov/ODOT/TD/TP/Reports/Value_of_TravelTime.pdf)

5. Applications and Extensions of Value of Travel-Time Theory to Travel in CAVs
-----------------------------------

* **Changing marginal utility of travel time** The theories for the value of travel time exposed in Small 2012 have the common feature that they all rely on travel time $T_T$ displacing other activities, i.e. other types of time use, through the time budget constraint $\Sigma_i^K T_i \le \bar T$$. Except for referring to the risk-burden/effort associated with congested travel time, they do not speak of the _quality_ of time spent while traveling.  So quality of travel-time must be embedded in the marginal utility of travel time $U_{TT}$.
* **Multiple dimensions (types or components) of travel time** Need to disaggregate travel time by mode (car, transit, foot/bike) and component (congested travel, recreational travel, local vs. highway or long-distance travel) or purpose (commute, shopping/errand, etc.)  Small (2012) and others document the heterogeneity of travel time, and CAV technology may affect the value of time in each of these activities differentially
* **Non-linearity of Travel Time Cost** There is some evidence that the utility of travel time is not linear in trip length.  How can this be included in the theory, and can we then consider the implications of connectivity/automation?
  * Small notes that Ben-Akiva and Lerman (1985) use a piece-wise linear specification for travel demand, and conclude that the first 20 minutes of a commute trip have little or no cost, presumable due to a high consumption benefit associated with spending a short time commuting.
* **Travel Time Reliability**

6. Ongoing/Initial VOTT Research for CAVs
--------------------

#### Cyganski et al 2014
* Rita Cyganski, Eva Fraedrich, Barbara Lenz 2014. TRAVEL-TIME VALUATION FOR AUTOMATED DRIVING: A USE-CASE-DRIVEN STUDY, Paper Submitted for presentation and publication at TRB 94th 27 Annual Meeting, 
http://elib.dlr.de/95260/1/TRB2015_CYGANSKI_FRAEDRICH_LENZ_15-4259.pdf
    *  quantitative online survey in Germany. The study includes detailed information on respondents’ current use of, and attitudes towards, today’s available transport modes, as well as their perception of the advantages of automated vehicles. Special focus is placed on current and future travel-time use
    * Perceived benefits with respect to 49 time use were mainly window gazing and relaxing, whereas the possibility of working 50 seemed to be less valued. A probit model analysis examined influencing factors on the 51 propensity of regarding working while traveling as an advantage - a strong significance of the 52 current time use on public transport and long-distance train trips, gender and rational attitude 53 towards public transport was revealed.

#### Kamin and Cody 201X (MIT)
* Kamin, Cody and Danial Morton 201X [Valuing the Convenience of Fully Autonomous Vehicles] (https://higherlogicdownload.s3.amazonaws.com/AUVSI/3a47c2f1-97a8-4fb7-8a39-56cba0733145/UploadedImages/documents/pdfs/Posters/Valuing%20the%20Convenience%20of%20Fully%20Autonomous%20Vehicles.pdf)

> In this work, we expand on the methodology for valuing time
used by transportation planners. We do this in two ways. 1) We
look beyond on-road travel time and account for the value of the
time saved parking and walking. 2) We account for the value of
being able to engage in work or leisure while traveling. The
resultant value of travel time savings (VTTS) is more than five
times larger the often cited safety benefits of this technology. ...
> Through a review of the relevant literature
and data sets we develop estimates for both
the amount of time spent on mobility-related
activities and the value of that time. In
particular, multiple comparisons are used to
develop a VTTS weighting for travel by SDC.
For each travel scenario, we then apply these
values to the USDOT formula for calculating
the value of travel time.
* Kamin and Morton compile estimates of time spent on travel-related activities like waiting, traveling, walking, parking, walking from vehicle, servicing and fueling. They estimate the changes in time spent on those activities and the value of time saved.
* They do _not_ appear to provide fundamental new estimates of the VOTT/VTTS while traveling in CAVs.

#### Zmud et al 2015, 2016
* This study gathered empirical [survey] evidence on adoption patterns of self-driving vehicles, people's likely use of them, and how that might influence amount of travel, mode choice, auto ownership, and other travel behavior decisions.
* Online survey followed by qualitative interviews, by Among those who indicated intent to use, conducted to ascertain the impact on their travel behavior

Mini-Biblio Summaries 
------------
* Cervero, Robert, 2001. "Induced Demand: An Urban and Metropolitan Perspective."" Prepared for Policy Forum: Working Together to Address Induced
Demand. Berkeley, CA. http://www.uctc.net/research/papers/648.pdf
    * Published as: "Induced Demand: An Urban and Metropolitan Perspective," in _Working Together to Address Induced Demand_, Washington, D.C., Eno Transportation Foundation, 2002, pp. 55-73.
    * Surveys 30 years of literature on Induced demand, which mostly focused on the elasticity of VMT to highway improvements or expansions (lane miles)
        * Of course there are problems with using lane-miles as a measure of highway capacity expansion
        * Since CAVs would reduce congestion and increase traffic through-put, their effect may be\ similar to an expansion of highway lanes. One important difference is that their effect can apply anywhere in the system, rather being localized to one highway segment like a roadway improvement.
        * Also, since the microeconomic theory indicates that road improvements alter travel demand through altering the intermediate variable of "Generalized Travel Cost" (by altering time and safety), in that sense, CAVs are similar to roadway improvements.  (NHTSA has known this for decades).
        * Cervero writes that the theory is: 
            * travel demand (number of trips) is a function of generalized cost $C_G$ and other control variables $x$, $Q_T = f(C_G, x)$
            * generalized cost is a function of road supply $S$ and other control variables $x$, $C_G = g(S, x)$
            * elasticity of travel demand with respect to generalized cost: $\eta_{T,GC} = d ln(Q)/ d ln(C_G)$
        * Cervero (2001:10) reports that virtually all empirical studies to date fail to properly represent these nested relationships between road improvements and induced travel.  They either omit the intermediate causal step of road capacity altering generalized cost, or "fail to weigh the relative role of road improvements in bringing about travel-time safings and to guage how in turn this induces travel."
    * some/many induced demand effects are actually shifts in route or timing of the trips, rather than new trips (latent trips, mode shifts, longer trips).  
        * Cervero finds "Most consistent across studies is the position that improved roads trigger shifts [in travel time] from the off-peak to the peak."


* Le Vine, S., Zolfaghari, A., & Polak, J. (2015). Autonomous cars: The tension between occupant experience and intersection capacity. Transportation Research Part C: Emerging Technologies, 52, 1-14. doi:10.1016/j.trc.2015.01.002
    * Non-drivers can engage in other activities, but experience discomfort at lower levels of acceleration than drivers.  Authors "found that restricting the dynamics of autonomous cars to the acceleration/deceleration characteristics of both rail systems [to improve ride quality leads to reductions in a signalized intersection's vehicle-processing capacity and increases in delay. The impacts were found to be larger when constraining the autonomous cars' dynamics to the more-restrictive acceleration/deceleration profile of high-speed rail."

* Lyons, G., & Urry, J. (2005). Travel time use in the information age. Transportation Research Part A: Policy and Practice, 39(2-3 SPEC. ISS.), 257-276. doi:10.1016/j.tra.2004.09.004
    * In transport scheme appraisal, savings in travel time typically represent a substantial proportion of the benefits of a scheme.... Such benefits are founded on the assumption that travel time is unproductive,
wasted time in-between "real" activities .... Travel demand analysis treats
travel time and activity time as separate, albeit acknowledging an interdependency.  The paper challenges these approaches by exploring how travel time can be, and is, being used 'productively' as activity time, .... Such undermining of the division between activities and travelling, and between activity time and travel
time, may have major implications for future levels of mobility, for the modal distribution of travel, for the validity of current transport appraisal methodology ....

* Ory, D. T., & Mokhtarian, P. L. (2005). When is getting there half the fun? Modeling the liking for travel. Transportation Research Part A: Policy and Practice, 39(2-3 SPEC. ISS.), 97-123. doi:10.1016/j.tra.2004.09.006
    * explores the notion of "directionless travel," i.e. travel for its own sake
    * eight models of short-distance Travel Liking and five models of long-distance
    * studied "Travel Liking"

* Steg, L. (2005). Car use: Lust and must. Instrumental, symbolic and affective motives for car use. Transportation Research Part A: Policy and Practice, 39(2-3 SPEC. ISS.), 147-162. doi:10.1016/j.tra.2004.07.001
    * This paper reports results of two questionnaire studies aimed at examining various motives for car use. ... As proposed by Dittmar's (1992) [The social psychology of material possessions: to have is to be.] model on the meaning of material possessions, results from both studies revealed that car
use not only fulfils instrumental functions, but also important symbolic and affective functions. Second, it was studied to what extent these different motives are related to the level of car use. ... it appeared that commuter car use was most strongly related to symbolic and affective motives, and not to instrumental motives. ... These results suggest that policy makers should not exclusively focus on instrumental motives for car
use, but they should consider the many social and affective motives as well.


* Sheikh, Adnan, Angshuman Guin, Randall Guensler 2014. Value of Travel Time Savings
Evidence from I-85 Express Lanes in Atlanta, Georgia, Transportation Research Record: Journal of the Transportation Research Board
http://trrjournalonline.trb.org/doi/pdf/10.3141/2470-17

> The I-85 express lanes in the Atlanta, Georgia, metropolitan region are among the newest high-occupancy-toll (HOT) facilities in the United States. This paper examines travelers' value of time and their willingness to pay for the I-85 express lanes, compares travel times and travel time reliability in the general purpose and HOT lanes, and discusses the benefits users gain from the facility. The study examined 9 months of toll and travel time data for drivers who traversed the entire length of the express lane corridor during the morning and afternoon peak periods. Median values of travel time savings were $36/h in the southbound morning peak period and $26/h in the northbound afternoon peak period. The express lanes exhibited reliability benefits relative to the general-purpose lanes, but this study does not quantify the value of those benefits because of the lack of stated-preference data. The value that HOT lane users attributed to their time saved exceeded the time value obtained by using the average wage rate in the region. These results for the value of travel time were found to be similar to those for HOT lane implementations in other locations.

* Athira, I. C., Muneera, C. P., Krishnamurthy, K., & Anjaneyulu, M. V. L. R. (2016). Estimation of Value of Travel Time for Work Trips. Transportation Research Procedia, 17(December 2014), 116–123. http://doi.org/10.1016/j.trpro.2016.11.067

> Fezzi et al (2012) estimated the value of travel time specific for recreation trips by revealed preference approach. The national value of time studies conducted in many countries (UK, Sweden, Finland, France etc.) considered the experimental data collection methods such as stated preference, along with results based on conventional revealed
preference method. Wardman (1987) conducted an empirical study using stated preference data to determine the distribution of individual values of time. Ahsen et al (2002), Richardson (2004) Antoniou et al (2007), Tseng and Verhoef (2008) and Xumei et al (2011) also employed stated preference approach for estimating VOT


* Small, Winston and Yan 2005

> Those based solely on data describing actual choices, i.e., revealed preference (RP) data, have been hampered by collinearity among cost and travel-time variables; consequently they rarely have accounted for reliability (i.e., the predictability of travel time) and they have not accounted for het- erogeneity in cost or travel-time elasticities arising from unobserved sources. ... [Other studies]
Measure the extent of unobserved heterogeneity using stated preference (SP) data that describe hypothetical responses, but SP data are tainted by doubt whether behavior exhibited in hypothetical situations applies to actual

* Mokhtarian, P., and I. Salomon, 2001. "How Derived is the Demand for Travel? Some Conceptual and Measurement Considerations" Transp. Res. A, 35(8), September 2001, pp. 695-719
http://repositories.cdlib.org/itsdavis/UCD-ITS-REP-01-15/
    * contests the conventional wisdom that travel is a derived demand "at least as an absolute"
    * suggests that under some circumstances, travel is desired for its own sake. 
    * calls the phenomenon "undirected travel," where travel is the activity itself, not a by product of the activity

* [Mohammadian2016 Research Results in New Mode Choice Model that Gives More Accurate Data](https://utc.uic.edu/new-mode-choice-model-gives-transportation-agencies-tool-to-make-more-informed-planning-decisions-news-story/) Abolfazl (Kouros) Mohammadianand Mahmoud Javanmardi, APR 21, 2016

> The modeling project is detailed in a report, “Mode Choice Modelling Using Personalized Travel Time and Cost Data,” produced by researchers affiliated with the Urban Transportation Center at the University of Illinois at Chicago. As stated in the report: “With a focus on transit, the study reveals how transit alternatives are limited/unavailable in certain areas which must be taken in to account in our mode choice models.”
Researchers analyzed more than 170,000 trips taken by people from 14,000 households who used auto, transit, cycling and walking modes of transportation. The travel data was collected through the CMAP Travel Tracker Survey conducted by the Chicago Metropolitan Agency for Planning (CMAP) from people within eight counties surrounding Chicago.
The model developed stands apart from other transportation models because researchers incorporated point-to-point data from Google and Bing maps for travel time information during all trips reported in travel survey. This data was used to make the model estimation more accurate.

* [Mohammadian 2016. Mode Choice Modelling Using Personalized Travel Time and Cost Data](https://utc.uic.edu/research/mode-choice-modelling-using-personalized-travel-time-and-cost-data/)

* Javanmardi et al. 2016:
* Javanmardi, M., Langerudi, M. F., Anbarani, R. S., & Mohammadian, A. (2016). Mode Choice Modelling Using Personalized Travel Time and Cost Data, (312). Retrieved from https://utc.uic.edu/wp-content/uploads/Mode-Choice-Modelling-Using-Personalized-Travel-Time-and-Cost-Data.pdf

> This paper describes the importance of accurate information by analyzing a travel survey and proposes a methodology to generate the actual alternatives that individuals had when making their trips using Google Maps API and RTA’s Goroo TripPlanner.

  	* Develop a "methodology to generate the actual alternatives that individuals had when making their trips using Google Maps API and RTA’s Goroo TripPlanner."
  	* More detailed trip option data such as this "reveals how some alternative modes’ travel information are limited\unavailable," which must be accounted mode choice models
  	* * Provides statistics regarding available alternatives and the constraints on travelers' choices
  	* "the purpose of this study is to describe the advantages and importance of accurate 11 data and resource constraints in the models within mode choice of activity-based frameworks"

    * Javanmardi et al. find that mode choice for successive trips in a tour (e.g. return home trips) are strongly constrained when a transit or auto-drive tour starts with either auto-drive or transit. This inability to switch modes may be dramatically altered by the introduction of AVs, or particularly SAVs, since there is not commitment to a particular mode on successive links.
    * "One of the analyses that were conducted was the distribution of modes, when previous mode of a trip was auto-drive. The [analysis] shows that almost all return-home trips and more than 96% of other trips use the same mode as their previous mode (i.e. auto-drive)."
    * "This represents a constraint that individuals encounter when choosing a mode and confirms the expectation that when a transit or auto-drive tour starts, the mode is typically fixed in the tour.""


* Mohammadian, Abolfazl (Kouros), & Javanmardi, M. (2016). Research Results in New Mode Choice Model that Gives More Accurate Data. Mass Transit (Online), April 21, 1–7. Retrieved from http://www.masstransitmag.com/article/12190856/research-results-in-new-mode-choice-model-that-gives-more-accurate-data
    * The model is described in the paper: "Mode Choice Modelling Using Personalized Travel Time and Cost Data.” Researchers under the direction of Civil and Materials
Engineering Professor Dr. Abolfazl (Kouros) Mohammadian
    * used Travel Tracker Survey data collected by the Chicago Metropolitan Agency for Planning (CMAP), the
state-managed metropolitan planning organization, to provide information from eight counties surrounding Chicago. The population sampled in the survey matched
the region’s household population by county. 
    * Designed for travel modeling, this information let researchers analyze more than 170,000 trips taken by people during a
14-month period starting in 2007. An estimated 218,000 travel activities were recorded.
    * what is not collected here is information of other modes of travel (e.g. how long it would it take to travel by bus or train).

    * Focused on trips originating from home: 
        > "since trips that are not home-based are highly associated with their respective home-based origin trip within a tour (a tour is a combination of trips, which starts and ends at home). This means when an individual
starts a tour which is comprised of several trips, the mode of the first trip that originates from home highly affects the mode of the later trips.”
    * Findings:
    * "Alternative modes of travel were not available to travelers for many trips studied. (lack of transit stops or stations close to the origin
    or destination, and a lack of transit service during the time of the trip)
    * importance of incorporating state-of-the-art data collected by companies similar to Google, NAVTEQ, and Microsoft for more effective transit planning models."


References
----------

* Ayala, R. (2014). Revised Departmental Guidance on Valuation of Travel Time in Economic Analysis. Retrieved from https://cms.dot.gov/sites/dot.gov/files/docs/USDOT VOT Guidance 2014.pdf

* Bansal, P., Kockelman, K. M., & Singh, A. (2016). Assessing public opinions of and interest in new vehicle technologies: An Austin perspective. Transportation Research Part C: Emerging Technologies, 67, 1-14. doi:10.1016/j.trc.2016.01.019

* Becker G (1965) A theory of the allocation of time. Econ J 75:493–517
http://www.jstor.org/stable/2228949

* Belenky, P. (2011). Revised Departmental Guidance on Valuation of Travel Time in Economic Analysis. Office of Transportation Policy Reports, 1–28. http://doi.org/10.1017/CBO9781107415324.004
https://www.transportation.gov/sites/dot.dev/files/docs/vot_guidance_092811c.pdf

* Ben-Akiva, M., Lerman, S.R., 1985. Discrete Choice Analysis: Theory and Application to Travel Demand. MIT Press, Cambridge, MA.

* Cervero, Robert, 2001. "Induced Demand: An Urban and Metropolitan Perspective."" Prepared for Policy Forum: Working Together to Address Induced
Demand. Berkeley, CA. http://www.uctc.net/research/papers/648.pdf
    * Published as: "Induced Demand: An Urban and Metropolitan Perspective," in _Working Together to Address Induced Demand_, Washington, D.C., Eno Transportation Foundation, 2002, pp. 55-73.

* Endorf, R., & U.S. Department of Transportation. (2015). Revised Departmental Guidance on Valuation of Travel Time in Economic Analysis. Retrieved from https://www.transportation.gov/resources/2015-revised-value-of-travel-time-guidance

* Fagnant, D. J., & Kockelman, K. (2015). Preparing a nation for autonomous vehicles: Opportunities, barriers and policy recommendations. Transportation Research Part A: Policy and Practice, 77, 167-181. doi:10.1016/j.tra.2015.04.003
    * Also published as Eno paper in 2013, https://www.enotrans.org/wp-content/uploads/2015/09/AV-paper.pdf

* Fagnant, D. J., & Kockelman, K. M. (2014). The travel and environmental implications of shared autonomous vehicles, using agent-based model scenarios. Transportation Research Part C: Emerging Technologies, 40, 1-13. doi:10.1016/j.trc.2013.12.001

* Gucwa, Michael, 2014. [Mobility and Energy Impacts of Automated Cars](https://higherlogicdownload.s3.amazonaws.com/AUVSI/c2a3ac12-b178-4f9c-a654-78576a33e081/UploadedImages/documents/pdfs/7-16-14%20AVS%20presentations/Michael%20Gucwa.pdf). Presentation to the Automated Vehicles Symposium. Burlingame, CA, July 16. 

* Guriau, M., Billot, R., El Faouzi, N. E., Monteil, J., Armetta, F., & Hassas, S. (2016). How to assess the benefits of connected vehicles? A simulation framework for the design of cooperative traffic management strategies. Transportation Research Part C: Emerging Technologies, 67, 266-279. doi:10.1016/j.trc.2016.01.020

* Hensher, David A., 1977. Value of Business Travel Time. Pergamon Press, Oxford.

* Hensher, David A., 2011. Valuation of travel time savings. In: de Palma, A., Lindsey, R., Quinet, E., Vickerman, R. (Eds.), A Handbook of Transport Economics.
Edward Elgar, pp. 135–159. (Chapter 7).

* Hess, S., Bierlaire, M., & Polak, J. W. (2005). Estimation of value of travel-time savings using mixed logit models. Transportation Research Part A: Policy and Practice, 39(2-3 SPEC. ISS.), 221-236. doi:10.1016/j.tra.2004.09.007

* ITF (2009), Value of Travel Time Reliability and Cost-Benefit Analysis, International Transport Forum (www.internationaltransportforum.org); Available at: www.internationaltransportforum.org/Proceedings/reliability/index.html.

* Jamson, A. H., Merat, N., Carsten, O. M. J., & Lai, F. C. H. (2013). Behavioural changes in drivers experiencing highly-automated vehicle control in varying traffic conditions. Transportation Research Part C: Emerging Technologies, 30, 116-125. doi:10.1016/j.trc.2013.02.008

* Le Vine, S., Zolfaghari, A., & Polak, J. (2015). Autonomous cars: The tension between occupant experience and intersection capacity. Transportation Research Part C: Emerging Technologies, 52, 1-14. doi:10.1016/j.trc.2015.01.002

* Litman, Todd (2009), "Travel Time," Transportation Cost and Benefit Analysis, Victoria Transport Policy Institute (www.vtpi.org). Available at: www.vtpi.org/tca/tca0502.pdf. 

* Lyons, G., & Urry, J. (2005). Travel time use in the information age. Transportation Research Part A: Policy and Practice, 39(2-3 SPEC. ISS.), 257-276. doi:10.1016/j.tra.2004.09.004

* Mersky, A. C., & Samaras, C. (2016). Fuel economy testing of autonomous vehicles. Transportation Research Part C: Emerging Technologies, 65, 31-48. doi:10.1016/j.trc.2016.01.001

* Javanmardi, M., Langerudi, M. F., Anbarani, R. S., & Mohammadian, A. (2016). Mode Choice Modelling Using Personalized Travel Time and Cost Data, (312). Retrieved from https://utc.uic.edu/wp-content/uploads/Mode-Choice-Modelling-Using-Personalized-Travel-Time-and-Cost-Data.pdf

* Mokhtarian, P. L. (2005). Travel as a desired end, not just a means. Transportation Research Part A: Policy and Practice, 39(2-3 SPEC. ISS.), 93-96. doi:10.1016/j.tra.2004.09.005

* Mokhtarian, P., and I. Salomon, 2001. "How Derived is the Demand for Travel? Some Conceptual and Measurement Considerations" Transp. Res. A, 35(8), September 2001, pp. 695-719
http://repositories.cdlib.org/itsdavis/UCD-ITS-REP-01-15/
    
* Mokhtarian, P. L. (2005). Travel as a desired end, not just a means. Transportation Research Part A: Policy and Practice, 39(2–3 SPEC. ISS.), 93–96. http://doi.org/10.1016/j.tra.2004.09.005

* Ory, D. T., & Mokhtarian, P. L. (2005). When is getting there half the fun? Modeling the liking for travel. Transportation Research Part A: Policy and Practice, 39(2–3 SPEC. ISS.), 97–123. http://doi.org/10.1016/j.tra.2004.09.006

* Pawlak, J., Polak, J. W., & Sivakumar, A. (2015). Towards a microeconomic framework for modelling the joint choice of activity-travel behaviour and ICT use. Transportation Research Part A: Policy and Practice, 76, 92-112. doi:10.1016/j.tra.2014.10.013

* Pickrell, D. (2001). Induced demand- its definition, measurement and significance. In Eno Transportation Foundation Policy Forum. Retrieved from http://ntl.bts.gov/lib/51000/51100/51199/InducedDemand.pdf

* Sheikh, Adnan, Angshuman Guin, Randall Guensler 2014. Value of Travel Time Savings
Evidence from I-85 Express Lanes in Atlanta, Georgia, Transportation Research Record: Journal of the Transportation Research Board
http://trrjournalonline.trb.org/doi/pdf/10.3141/2470-17

* Small, Kenneth A. (2012). Valuation of travel time. Economics of Transportation, 1(1-2), 2-14. doi:10.1016/j.ecotra.2012.09.002

* TRISP (2005), “Valuation of Travel Time Saving,” Economic Evaluation Notes, UK Department for International Development and the World Bank (www.worldbank.org); at http://go.worldbank.org/ME49C4XOH0. Summarizes travel time valuation methods suitable for developing country transport project evaluation.

* Train, Kenneth 2003.  Discrete Choice Methods with Simulation, (Cambridge University Press).

* U.S. Department of Transportation (2003), Revised Departmental Guidance: Valuation of Travel Time in Economic Analysis, Office of the Secretary of Transportation, U.S. Department of Transportation. Available at: http://ostpxweb.dot.gov/policy/Data/VOTrevision1_2-11-03.pdf.

* VTPI2017 Travel Time Costs
http://www.vtpi.org/tca/tca0502.pdf

* Wardman, M., & Lyons, G. (2015). The digital revolution and worthwhile use of travel time: implications for appraisal and forecasting. Transportation, 507-530. doi:10.1007/s11116-015-9587-0

* White, V., & U.S. Department of Transportation. (2016). Revised Departmental Guidance on Valuation of Travel Time in Economic Analysis. Retrieved from https://www.transportation.gov/office-policy/transportation-policy/revised-departmental-guidance-valuation-travel-time-economic

* Zamparini Luca and Aura Reggiani (2007), “Meta-Analysis and the Value of Travel Time Savings: A Transatlantic Perspective in Passenger Transport,” Networks and Spatial Economics (www.springerlink.com); at www.springerlink.com/content/h663q51u448078x1. 


* Zmud, J., Sener, I. N., & Wagner, J. (2016). Consumer Acceptance and Travel Behavior Impacts of Automated Vehicles Final Report, Texas A&M Transportation Institute PRC 15-49 F. Retrieved from http://tti.tamu.edu/documents/PRC-15-49-F.pdf







Quo Vado?
------------

```{r testBlock}
# commented code
x = rnorm(1000)
```

