---
title: "SAV Replacement of Private Household Conventional Vehicles and Implications for Vehicle Fleet Size"
author: "Paul N. Leiby"
date: "3/18/2019 - draft"
output:
  word_document: default
  html_document: default
---

leibypn@ornl.gov
Oak Ridge National Laboratory

This note summarizes estimates of the number of private (household) conventional vehicles that could be replaced by each Shared Automated Vehicle (SAV) in a regional transportation system, essentially providing the same service level in terms of ride wait/travel time. We call this the "SAV Replacement Ratio," and it depends on the domain of vehicle operation (region, road network and time of day), depends strongly on the extent of ride-pooling, and to some extent on vehicle technology. Estimates are entirely from various published simulation studies, with some improving fidelity over time. The second part of this note shows simple calculations of how the total vehicle fleet (parc) size could decline, if the SAVs introduced replace the rides provided by multiple private cars (according to the SAV Replacement Rate), while the total number of rides remains constant.

### 1. Estimates of SAV Replacement Ratio in the Literature

Results for the "SAV Replacement Ratio" (number of private conventional household vehicles that could be replaced by commercially-operated SAVs while providing roughly equivalent service times) necessarily vary with assumed system conditions. They depend on:

- service area attributes, such as the density of travel activity and the type of road network on which vehicles operate;
    - e.g. studies may not include full metro areas, and ride hailing/pooling expected to be near-zero in rural area travels (~20% of total trips). 
- simulation algorithms (dispatch and routing heuristics)
- vehicle drivetrain and fueling/charging infrastructure (determining out-of-service time for recharging)
- acceptable wait times (1 to 5 minutes typically assumed)
- traffic congestion (related to time of operation)
- extent of ride-pooling (Dynamic Ride Sharing, or DRS)
- assumed extent of replacement of personal trips, in the region with SAV trips. 
    - Many extant studies exogenously assume some fraction of trips will be taken on SAVs, and ask how many SAVs would be needed, rather than model the fraction of travel on SAVs, reflecting consumer choice).

Comment: It appears that essentially _all_ estimates involve a central or global system optimization, i.e. equivalent to a _single_ coordinated ride-service operator or TNC, seeking to provide acceptable wait and travel times with minimal SAV fleet size.^[Thus they may imply either a single coordinated SAV fleet under city/government control or a utility-like arrangement where a service monopoly is granted to, or achieved by, a single operator. Perhaps more realistically, in a competitive market with potentially substantial profits, we would expect the entrance of multiple competing ride services, and lower levels of SAV utilization until costs and profits equilibrate. This would imply lower SAV Replacement Ratio] These estimates are generally demand driven.


**Table 1: Estimates of the Replacement Ratio - Household Vehicles to SAVs Needed to Satisfy Demand**

**Private Veh: SAV** | **Notes**                       | **Source**
-------------------|------------------------------|------------
3:1                   |Singapore                    | Speiser et al 2014.
9:1                   |Early widely-cited simulation of Austin, 12x24 mi. core area. No pooling, congestion|Fagnant & Kockelman, 2014, 2016
7:1 - 11:1            |Range of simulations in urban (Austin) area, coordinated optimum depends on road network, acceptable (short) wait times for pick-up. Replacement higher for grid networks. These studies typically *omit/simplify congestion effects*, assume SAVs completely replace personal vehicles.| (Fagnant & Kockelman, 2014, 2016, Chen, Kockelman, & Hanna, 2016, Loeb, Kockelman, & Liu, 2017)
17:1                  |Pooled SAVs. Little explanation given. | Johnson 2016.
3.6:1                 |Peak travel hour, no DRS/pooling (congested, similar origins and destinations in rush hour)|Levin et al. 2017
31:1                  |Peak estimate in Central business district, extensive DRS/pooling (pooling of 3-4 passengers) |Levin et al. 2017


### 2. Estimating Vehicle Parc Size for Levels Replacement Ratio, and SAV Fraction

We show simple calculations of how the total vehicle fleet (parc) size could decline, if 

- the SAVs introduced replace rides provided by multiple private cars (according to the SAV Replacement Rate),
- the utilization rate (rides per car) for private household vehicles remains constant,
- while the total number of rides remains constant.

#### Base Case - All Private Household Vehicles

Consider the size of the vehicle fleet, or parc, in the base case (Case 0) $N_{tot0}$. Assuming for simplicity that  in the Base Case all vehicles are private (and manually driven?), the number of Private vehicles $N_{PV0}$ equals the total number of vehicles:

$$ParcSize_0 \equiv N_{tot0} = N_{PV0}$$

For convenience and generality to any transport region and time interval, we normalize the base number of rides per private household vehicle to 1.0.^[More specifically, the average number of rides (roadway trips) provided per household vehicle per year in the U.S. is XXX, while in a specific urban area and hour of the day it may be YYY.] Note that with this normalization, the number of rides provided per SAV is the same as the relative number of ride, i.e. equal to the Replacement Ratio $RR$.^[This is also the approach and interpretation of shared vehicle activity in, e.g. Levin et al. 2017]

This base case number of household vehicles satisfies the base demand for rides/trips, $Rides_0$.
$$Rides_0 = 1 \cdot N_{tot0}$$

#### Alternative Case with SAVs

In the alternate case (Case 1), there are $N_{PV1}$ private vehicles, $N_{SV1}$ shared vehicles, 
and shared vehicles have a *Replacement Ratio* of $RR$, 
meaning each shared vehicle provides $RR$ as many rides as a private vehicle. 

We are interested in the relative total fleet/parc size compared the base case, given a replacement ratio
$RR$ and a given fraction i=$S_{SV1}$ of vehicles that are SAVs.

For $\rho$ denoting the Relative Vehicle Parc Size, i.e.
$$\rho \equiv N_{tot1}/N_{tot0} = (N_{PV1} + N_{SV1})/N_{PV0}$$
We assume the total rides provided (total travel demand) in the alternate case are the same as in the base case.
$$Rides_1 = Rides_0$$
If the utilization rate of private vehicles is unchanged, i.e. still 1.0, this means.
$$Rides_1 = 1 \cdot N_{PV1} + RR \cdot N_{SV1} = Rides_0 = 1 \cdot N_{PV0}$$
So the equal ride service condition implies a relationship between the base and alternate vehicle numbers:
$$1 \cdot N_{PV1} + RR \cdot N_{SV1} = N_{PV0}$$
Looking at the fleet/parc size in the alternative case
$$ParcSize_1 \equiv N_{tot1} = N_{PV1} + N_{SV1} $$

Looking at the Relative Vehicle Parc Size
$$\rho \equiv \frac {N_{tot1}} {N_{tot0}} = \frac {N_{PV1} + N_{SV1}} {N_{PV0}} = \frac {N_{PV1} + N_{SV1}} {1 \cdot N_{PV1} + RR \cdot N_{SV1}}$$

or
$$\rho = \frac {N_{PV1} + N_{SV1}}{1 \cdot N_{PV1} + RR \cdot N_{SV1}}$$
Dividing both the numerator and denomnator by the total number of vehicles in the alternative case
($N_{tot1} \equiv N_{PV1} + N_{SV1}$) allows us the express the relative parc size in terms of the replacement ratio and the
assumed fraction of vehicles that are SAVs.

$$\rho = \frac {1}{(N_{PV1}/N_{tot1} + RR \cdot N_{SV1}/N_{tot1})}$$


So for any posited fraction $S_{s}$ of shared vehicles out of total vehicles $N$
$$S_{s1} == N_{SV1}/N_{tot1}$$
$$1-S_{s1} == N_{PV1}/N_{tot1}$$

$$\rho = \frac {1}{(1-S_{s1}) + RR \cdot S_{s1}}$$

Check:

Note that this algebraic result checks out, in that we can compute the number of rides provided by this number of vehicles:
$$Rides_1 = 1 \cdot N_{PV1} + RR  \cdot  N_{SV1}$$
$$\begin {align}
Rides_1 = N_{tot1}  \cdot  [1 \cdot (1-S_{s1}) + RR  \cdot  S_{s1}]\\
= N_{tot1}  \cdot  [(1-S_{s1}) + RR  \cdot  S_{s1}]
\end {align}$$

### Examples of Total Fleet/Parc Size for Different SAV Replacement Ratios and Shares of the Fleet
```{r echo=F, message=F, warning=F}
library(tidyverse)
echoWorking = F
```

```{r calcAndPlotRelativeParcSize, echo=echoWorking}
RRvals = 1:20
SSvals = seq(from=0, to =1.0, length.out = 21)
parc = data.frame(matrix(nrow=length(RRvals), ncol=length(SSvals)))
#RRvals = data.frame(RR = 1:20)
#SSvals = data.frame(SS = seq(from=0, to =1.0, length.out = 20))
#x = full_join(RRvals, SSvals, by=NULL)

rparc <- function(rr, ss) {
  # compute relative vehicle parc size for Replacement Ratio rr, SAV Share ss
  return( 1/((1-ss) + rr * ss) )
}

for (r in seq_along(RRvals)) {
  for (c in seq_along(SSvals)) {
    parc[r,c] = rparc(RRvals[r], SSvals[c])
  }
}

names(parc) = c(paste0("S_",SSvals))
parc$RR = RRvals

x = parc %>% gather(key = "SS", value="rho", -RR) %>%
  mutate(SS = as.numeric(gsub(pattern = "S_", replacement ="", SS))) %>%
  arrange(RR)

x %>%  mutate(RR = as.factor(RR)) %>% 
  group_by(RR) %>%
  ggplot(aes(x=SS, y =rho, color=RR)) + geom_line() +
  ggtitle("Relative Fleet/Parc Size for Alternative SAV Replacement Ratios") + 
  xlab("SAV Share of Fleet") + ylab("Relative Fleet Size")
```


### Potential Extension: More Vehicle Types; Different Utilization Rates; Changed Total Travel Demand

This is an idealized calculation for two vehicle types, 
comparing the ride service provisions of SAVs with Private/Household Manual Vehicles. 

Potential extensions that would influence the estimate Replacement Ratio and vehicle Parc Size include:
- Differentiating between the Replacement Ratio for SAVs and SMVs (Shared Manual Vehicles may not serve as many rides);
- Differentiating between the Replacement Ratio for SAVs and Private AVs (which may still serve more rides than Private MVs, e.g. within a single household);
- Accounting for declining utilization of private household vehicles, without their scrappage/resale; qnd
- Accounting to changes in trips/VMT 

References
-----------------
- Chen, T. D., Kockelman, K. M., & Hanna, J. P. (2016). Operations of a shared, autonomous, electric vehicle fleet: Implications of vehicle & charging infrastructure decisions. Transportation Research Part A: Policy and Practice, 94, 243–254. http://doi.org/10.1016/j.tra.2016.08.020
- Fagnant, D. J., & Kockelman, K. M. (2014). The travel and environmental implications of
shared autonomous vehicles, using agent-based model scenarios. Transportation
Research Part C: Emerging Technologies, 40, 1–13.
- Fagnant, D. J., & Kockelman, K. M. (2016). Dynamic ride-sharing and fleet sizing for a
system of shared autonomous vehicles in Austin, Texas. Transportation, 1–16.
- Fagnant, D. J., Kockelman, K. M., & Bansal, P. (2015). Operations of a shared autonomous
vehicle fleet for the Austin, Texas market. Transportation Research, 28,
- Fagnant, D. J., & Kockelman, K. M. (2018). Dynamic ride-sharing and fleet sizing for a system of shared autonomous vehicles in Austin, Texas. Transportation, 45(1), 143–158. http://doi.org/10.1007/s11116-016-9729-z
- Johnson, Brian A. (Barclays Capital, 2016. Disruptive Mobility, Society of Automotive Analysts 2016 Outlook Conference 10 January 2016. https://societyofautomotiveanalysts.wildapricot.org/resources/Documents/Brian%20Johnson%20slides%20for%20Outlook.pdf
- Levin, M. W., Kockelman, K. M., Boyles, S. D., & Li, T. (2017). A general framework for modeling shared autonomous vehicles with dynamic network-loading and dynamic ride-sharing application. Computers, Environment and Urban Systems, 64, 373–383. http://doi.org/10.1016/j.compenvurbsys.2017.04.006
- Loeb, B., Kockelman, K.M., Liu, J., 2018. Shared autonomous electric vehicle (SAEV) operations across the Austin, Texas network with charging infrastructure decisions. Transp. Res. Part C Emerg. Technol. 89, 222–233. https://doi.org/10.1016/j.trc.2018.01.019
- Shaheen, Susan and Elliot Martin, 2019. [The Impact of Carsharing on Household Vehicle Ownership](http://www.accessmagazine.org/spring-2011/impact-carsharing-household-vehicle-ownership/).
- Segal, M., and Kockelman, K. M. 2017. Design and implementation of a shared
autonomous vehicle 1 system for Austin, Texas 2. In Kockelman, K. & Boyles, S.
(Eds.) Transportation research board 95th annual meeting, no. 16-1837
- Spieser, K., Treleaven, K., Zhang, R., Frazzoli, E., Morton, D., & Pavone, M. 2014.
Toward a systematic approach to the design and evaluation of automated mobility-
on-demand systems: A case study in Singapore. _Road vehicle automation_, 2014 229–245).



-------------------------------------------

Bibliography & Notes - Literature on CAVs SAVs Taxation and VMT
=========================

* See "SMART Mobility WTF  Scenarios - Ownership 20181218a.pptx"
Total Vehicle Stock Would Decline As Fraction of Shared Vehicles Rises

- 7:1 Austin six-county region in Austin, Texas (Loeb et al., 2017).
- DRS Fagnant and
Kockelman (2016) and Segal and Kockelman (2017) incorporated
dynamic ride-sharing, only 10% of Austin personal trips.

- Burns et al. (2013) studied a centrally dispatched SAV
system in three different urban and suburban environments. Their
findings indicated that a much smaller fleet of SAVs could provide
service to all residents with acceptable waiting times… Spieser et al. (2014) found
that a SAV fleet one-third the size of the personal vehicle fleet was
sufficient for providing service to Singapore travelers.


* Ana T. Moreno, Andrzej Michalski, Carlos Llorca, and Rolf Moeckel 2018. Shared Autonomous Vehicles Effect on Vehicle-Km Traveled and Average Trip Duration
_Journal of Advanced Transportation_
Volume 2018, Article ID 8969353, 10 pagesa
https://doi.org/10.1155/2018/8969353
Research Article
https://www.hindawi.com/journals/jat/2018/8969353/

* Johnson, Brian A. 2015. "Disruptive Mobility: AV Deployment Risks and Possibilities", Barklays Capital presentation, https://orfe.princeton.edu/~alaink/SmartDrivingCars/PDFs/Brian_Johnson_DisruptiveMobility.072015.pdf
    - SAV: 8% increase in VMT from empty trips, vehicle replacement ratio 9 (traditional vehicle per SAV), vehicle usage intensity x5. Pooled SAVs (PSAV): vehicle replacement ratio 15-18, VMT reduced by 40-50%, vehicle usage intensity x5, Car parc falls by 60% and annual LDV sales by 40% (SAAR, seaonall adjusted annual rate. Barclays calc.)
  - Discusses 4 alternative types of AVs: 
        1. Some traditional vehicles "for work and rural use," limited self-driving capabilities
            - rural areas are 24% of vehicles in use (NHTS)
        2. Family AV (FAVs) 2.1/HH declines to 1.2, mi/y incr. from 12k to 24k
        3. SAVs 9:1 replacement, 8% additional VMT empty miles, mi/y incr. from 12k to 64k
            - cost $0.64/mi per sedan SAV, $0.16/mi 2-seater
        4. PSAVs Pooled-SAVs: 15:1-18:1 replacement, 40-50% reduced VMT from shared rides, mi/yincr. from 12K to 64k
            - cost $0.21/mi-consumer sedan PSAV; $0.08/mile-consumer 2-seater PSAV

* Johnson, Brian A. (Barclays Capital, 2016. Disruptive Mobility, Society of Automotive Analysts 2016 Outlook Conference 10 January 2016. https://societyofautomotiveanalysts.wildapricot.org/resources/Documents/Brian%20Johnson%20slides%20for%20Outlook.pdf
  - Update of 2015 presentation, refined numbers. Slightly lower replacement rate, higher deadhead ZOV VMT, lower cost/mi
  - Discusses 4 alternative types of AVs: 
        1. Some traditional vehicles "for work and rural use," limited self-driving capabilities
            - rural areas are ~20% of vehicles in use (NHTS,  Kim, Jinwon & Brownstone, David)
“The impact of residential density on vehicle usage and fuel consumption”, 01/2010)
        2. Family AV (FAVs) 2.1/HH declines to 1.2, mi/veh-y incr. from 11346 to 22692 
        3. SAVs 7:1 replacement, 12% additional VMT empty miles, mi/veh-y incr. from 11346 to 64k
            - cost $0.29/mi to consumer per sedan SAV, $0.19/mi 2-seater
        4. PSAVs Pooled-SAVs: 17:1 replacement, 50%+ reduced VMT from shared rides, mi/yincr. from 11346 to 64k
            - cost $0.12/mi-consumer sedan PSAV; $0.10/mile-consumer 2-seater PSAV (no clear source)
                - "perpetual ride with ave wait time of 5 min"
    - Car parc falls by 60% and annual LDV sales (SAAR) by 40%, mass market by 70% Barclays calc.)
    - vehicle usage intensity (mi/yr) 5.6X for SAV and Pooled-SAV

* briefing.com 2019 Auto Sales: Light Vehicle Sales (SAAR). Auto Sales Data: Graphs of Monthly SAAR, Domestic and Total, Mar 2003 - Mar 2019, Source is Commerce Department.
https://www.briefing.com/Investor/Calendars/Economic/Releases/auto.htm?mod=article_inline

*  Fagnant, Kockelman, Bansal, “Operations of a Shared Autonomous Vehicle,” Transportation Research Record, No. 2536: 98-106, 2015.
> "Results show that each SAV is able to replace around 9 conventional vehicles within [service area] while still maintaining a reasonable level of service (as proxied by user wait times,
which average just 1.0 minutes)." Approximately 8% more VMT generated, unoccupied travel to the next traveler or repositioning.

* Fagnant, D. J., & Kockelman, K. M. (2018). Dynamic ride-sharing and fleet sizing for a system of shared autonomous vehicles in Austin, Texas. Transportation, 45(1), 143–158. http://doi.org/10.1007/s11116-016-9729-z

* Chen, T. D., Kockelman, K. M., & Hanna, J. P. (2016). Operations of a shared, autonomous, electric vehicle fleet: Implications of vehicle & charging infrastructure decisions. Transportation Research Part A: Policy and Practice, 94, 243–254. http://doi.org/10.1016/j.tra.2016.08.020

* Levin, M. W., Kockelman, K. M., Boyles, S. D., & Li, T. (2017). A general framework for modeling shared autonomous vehicles with dynamic network-loading and dynamic ride-sharing application. Computers, Environment and Urban Systems, 64, 373–383. http://doi.org/10.1016/j.compenvurbsys.2017.04.006

* Levin, M. W., Kockelman, K. M., Boyles, S. D., & Li, T. (2017). A general framework for modeling shared autonomous vehicles with dynamic network-loading and dynamic ride-sharing application. Computers, Environment and Urban Systems, 64, 373–383. http://doi.org/10.1016/j.compenvurbsys.2017.04.006

> 2.2. Shared autonomous vehicles
Multiple studies have investigated the possibility of using a
fleet of SAVs to reduce reliance on personal vehicles and improve
mobility and safety (Fagnant & Kockelman, 2015). Previous work
(Chen, Kockelman, & Hanna, 2016; Fagnant & Kockelman, 2014;
Loeb, Kockelman, & Liu, 2017) estimated that one SAV could replace
between seven and eleven private vehicles, depending on the network,
with small waiting times for pick-up. Higher replacement
rates were observed on grid networks (Chen et al., 2016; Fagnant
& Kockelman, 2014), whereas a rate of 1:7 was observed for a
six-county region in Austin, Texas (Loeb et al., 2017). Fagnant and
Kockelman (2016) and Segal and Kockelman (2017) incorporated
dynamic ride-sharing, and found that it could offset the additional
vehicle travel time. However, only 10% of personal trips of Austin
were included. Further studies on different cities have supported
indications that a smaller fleet of SAVs could provide service to
all travelers. Burns et al. (2013) studied a centrally dispatched SAV
system in three different urban and suburban environments. Their
findings indicated that a much smaller fleet of SAVs could provide
service to all residents with acceptable waiting times… Spieser et al. (2014) found
that a SAV fleet one-third the size of the personal vehicle fleet was
sufficient for providing service to Singapore travelers.


> ... This AM peak-hour scenario required far more SAVs than 1 per
9.3 travelers (Fagnant et al., 2015). One SAV could replace at most
3.6 personal vehicles because of congestion and peak-hour travel
patterns. Peak-hour travel is particularly difficult for SAV service

* Ashley Langer, Vikram Maheshr, Clifford Winston, 2017. From gallons to miles: A disaggregate analysis of automobile travel and externality taxes, _Journal of Public Economics_, 152, August 2017, pp 34-46

> Abstract
Policymakers have prioritized increasing highway revenues as rising fuel economy and a fixed federal gasoline tax have led to highway funding deficits. We use a novel disaggregate sample of motorists to estimate the effect of the price of a vehicle mile traveled on VMT, and we provide the first national assessment of VMT and gasoline taxes that are designed to raise a given amount of revenue. We find that a VMT tax dominates a gasoline tax on efficiency, distributional, and political grounds when policymakers enact independent fuel economy policies and when the VMT tax is differentiated with externalities imposed per mile.
https://www.sciencedirect.com/science/article/abs/pii/S0047272717300798 of 

* B. Starr McMullen, Lei Zhang, Kyle Nakahara, 2010. Distributional impacts of changing from a gasoline tax to a vehicle-mile tax for light vehicles: A case study of Oregon, a_Transport Policy_, 17(6)359-366,

> A vehicle-miles traveled (VMT) tax is frequently mentioned as viable alternative to a fuel tax for
collecting highway users fees from light vehicles. Both a static model and a regression based model are used here to assess the distributional impacts of a switch from a fuel tax to a VMT tax for the state of Oregon. The VMT tax is found to be slightly more regressive than the fuel tax and rural households are found to actually benefit relative to urban households under a VMT tax. Two alternative VMT structures that might increase incentives to use more fuel efficient vehicles are provided, but both are found to be even more regressive than a flat VMT tax.

* Simoni Et Al. Congestion pricing in a world of self-driving vehicles: An analysis of different strategies in alternative future scenarios

* Pettigrew, Dana and Norman, 2019. Clusters of potential autonomous vehicles users according to propensity to use individual vs shared vehicles - _Transport Policy_, 76, April 2019, pp. 13-20

* Gkartzonikas  Gritza 2019. What have we learned A review of stated preference and choice studies on autonomous vehicles - Trans Research - C.

> This paper provides a review of studies published in peer-reviewed journals, conference pro-ceedings, and technical academic and private sector reports on surveys about autonomous ve-hicles (AVs) from 2012 onward. The studies and respective surveys are categorized in this paperbased on the study objectives and methodology applied. More than half of the reviewed studieson AVs focus on capturing individuals’ behavioral characteristics and perceptions. The secondmost prevalent category includes studies about individuals’ willingness to pay to use AVs. Thereviewed studies were also categorized according to the study population. The paper identifiesand classifies attitudinal questions in each survey into different components that may affectbehavioral intention to ride in AVs and provides information on specific hypotheses that were setin the studies. Moreover, a discussion of the benefits, barriers/concerns, and opportunities re-lated to the deployment of AVs is presented. The paper concludes by summarizing the lessons learned and outlining the research gaps


#### Papers re TNC Empty vkt

Cramer, J., & Krueger, A. B. (2016). Disruptive Change in the Taxi Business: The Case of Uber. American Economic Review, 106(5), 177–182. http://doi.org/10.1257/aer.p20161002

Henao, A., & Marshall, W. E. (2018). The impact of ride-hailing on vehicle miles traveled. Transportation, (0123456789). http://doi.org/10.1007/s11116-018-9923-2

Lavieri, P. S., Dias, F. F., Juri, N. R., Kuhr, J., & Bhat, C. R. (2018). A Model of Ridesourcing Demand Generation and Distribution. Transportation Research Record. http://doi.org/10.1177/0361198118756628

Loeb, B., Kockelman, K. M., & Liu, J. (2018). Shared autonomous electric vehicle (SAEV) operations across the Austin, Texas network with charging infrastructure decisions. Transportation Research Part C: Emerging Technologies, 89(May 2017), 222–233. http://doi.org/10.1016/j.trc.2018.01.019

Schaller, B. (2017). Empty Seats, Full Streets. Fixing Manhattan’s Traffic Problem. Retrieved from http://schallerconsult.com/rideservices/emptyseats.pdf


Schaller, B., 2017. Empty Seats, Full Streets. Fixing Manhattan’s Traffic Problem, Schaller Consulting, December.
